---
title: "HW9"
author: "Group 12 - Shir, Omer & Ido"
date: "16/06/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

------------------------------------------------------------------------

### Paths and Data

```{r setup, message=FALSE, warning=FALSE}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }
if (!require('MASS'))       { install.packages('MASS');       library('MASS') }

source("../utils.R")
source("../lab5/lab5_utils.R")

healthy_file <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
chr1_file    <- "../lab1/data/chr1_str.rda"
```

```{r load-data, cache=TRUE}
healthy_reads <- fread(healthy_file, col.names = c("Chr", "Loc", "Length"))
load(chr1_file)
```

```{r bin-and-correct, cache=TRUE}
BIN <- 2500

# Two arms around centromere (120-130 Mb excluded)
dfA <- bin_data(healthy_reads, chr1, 5000001L,   119990000L, BIN)
dfB <- bin_data(healthy_reads, chr1, 130000001L, 242000000L, BIN)

add_pos <- function(df, beg) {
  df$pos <- beg + (seq_len(nrow(df)) - 1) * BIN + BIN / 2
  df
}
dfA <- add_pos(dfA, 5000001L)
dfB <- add_pos(dfB, 130000001L)
df  <- rbind(dfA, dfB)

# GC spline + plug-in estimator (same as lab 8 method b)
eps  <- 1
nz   <- df$reads > 0
fit  <- fit_gc_spline(df[nz, ])
fhat <- pmax(predict(fit, data.frame(gc = df$gc)), 0)
raw  <- (df$reads + eps) / (fhat + eps)
df$cn <- raw / median(raw[nz])

cat(sprintf("Total bins: %d  |  non-zero: %d\n", nrow(df), sum(nz)))
```

------------------------------------------------------------------------

## Part A -- Test Statistic

We need a statistic $T$ that summarises whether a genomic region has more or fewer reads than expected under copy number 1, while being **resistant to single noisy bins** -- meaning one extreme bin should not be able to drive the entire region to significance on its own.

We work on $\log(\hat{a}_k)$, the log of the GC-corrected copy-number estimate from Lab 8 (method b, single-sample spline correction).
Taking the log has two advantages: under $H_0$ (CN = 1) the statistic is centred at 0, and the distribution is approximately symmetric (raw $\hat{a}_k$ is right-skewed since CN cannot go below 0).
Gains produce positive values; losses produce negative values.

We aggregate over a sliding window of $W = 11$ bins ($\approx 27.5$ kb) and compare three candidates:

| Statistic | Formula | Robustness |
|----|----|----|
| **S1**: plain log | $\log(\hat{a}_k)$ -- per single bin, no aggregation | None -- baseline |
| **S2**: trimmed mean | $\overline{\log(\hat{a})}_{10\%}$ over window | Hard-discards top/bottom 10% of bins |
| **S3**: Huber location | $\hat{\mu}_\text{Huber}(\log(\hat{a}),\, k=1.345)$ over window | Continuously downweights outlier bins |

S1 serves as a baseline to show what robustness buys us.
Both S2 and S3 satisfy the requirement: no single bin can dominate the window-level statistic.
S3 (Huber) is the preferred choice because at $k = 1.345$ it achieves 95% efficiency under normality -- no power is lost when the data are well-behaved -- while degrading gracefully otherwise.
S2's hard trim boundary discards information for bins near the cutoff.

```{r define-statistics}
W    <- 11
HALF <- (W - 1) / 2

log_cn <- log(pmax(df$cn, 1e-6))   # log(a-hat), floored to avoid -Inf

# Rolling window helper: applies FUN to each bin's W-neighbourhood
roll_stat <- function(x, fun) {
  out <- rep(NA_real_, length(x))
  for (i in seq_along(x)) {
    idx <- max(1, i - HALF) : min(length(x), i + HALF)
    sub <- x[idx][df$reads[idx] > 0]   # only non-zero bins in window
    if (length(sub) >= 5) out[i] <- fun(sub)
  }
  out
}

df$S1 <- log_cn                                                  # plain log(a-hat)
df$S2 <- roll_stat(log_cn, function(x) mean(x, trim = 0.10))    # trimmed mean
df$S3 <- roll_stat(log_cn, function(x) huber(x, k = 1.345)$mu) # Huber
```

```{r stat-distributions, fig.width=11, fig.height=4, dev="png", dpi=180}
par(mfrow = c(1, 3), mar = c(4, 3.5, 2.5, 1))
stats <- list(
  list(v = df$S1[nz], lab = "S1: log(a-hat)  (per bin)"),
  list(v = df$S2[nz & !is.na(df$S2)], lab = "S2: trimmed mean of log(a-hat)"),
  list(v = df$S3[nz & !is.na(df$S3)], lab = "S3: Huber of log(a-hat)")
)
cols <- c("#555555", "#56B4E9", "#009E73")

for (i in seq_along(stats)) {
  v <- stats[[i]]$v
  v <- pmin(pmax(v, -1), 1.5)
  hist(v, breaks = 100, freq = FALSE, col = adjustcolor(cols[i], 0.6), border = NA,
       main = stats[[i]]$lab, xlab = "Statistic value", cex.main = 0.85)
  abline(v = 0, col = "red", lwd = 2, lty = 2)
  legend("topright", bty = "n", cex = 0.8,
         legend = sprintf("SD=%.3f\nMAD=%.3f", sd(v, na.rm=TRUE), mad(v, na.rm=TRUE)))
}
```

S2 and S3 are visibly tighter than S1 (lower SD/MAD) while remaining centred at 0 -- robustness without bias.

------------------------------------------------------------------------

## Part B -- Negative Control Regions and Null Distribution

**How we select the negative controls.**
We need genomic regions where CN = 1 with high confidence, so the statistic values there serve as our empirical $H_0$ distribution.
We use the paired healthy tissue sample (TCGA-13-0723-10B): a normal blood cell should have CN = 1 almost everywhere on chr1, making the healthy genome our natural candidate pool.
We compute the GC-corrected estimator on the healthy sample and filter for bins that are genuinely undisturbed, comparing two strategies:

- **Strategy A -- global MAD filter:** Keep every non-zero bin whose GC-corrected CN is within 3 MAD of 1. A loose global threshold; ~91% of bins pass. It removes only extreme outliers; everything else is treated as a control.

- **Strategy B -- local stability filter:** Keep a bin only if its 11-bin neighbourhood simultaneously has a median near 0 in log-space ($|\tilde{T}| \leq 0.095$) **and** a low local SD ($\leq 0.14$). Stricter (~28% of bins pass): a bin is accepted only if the whole local region is quiet. Bins at a CN-event boundary -- even if their own value is near 1 -- are excluded because their neighbours reveal the disruption.

**How we model $H_0$.**
We do not assume a parametric form for $H_0$.
The statistic values from the selected negative control bins *are* the empirical null distribution -- we treat them as a direct sample from the distribution of $T$ under the null hypothesis.
The histograms and QQ plots below characterise this distribution for each combination of statistic and strategy.
The key diagnostic is normality: a linear QQ plot means that both empirical and parametric (z-score) p-values will agree, which is a useful consistency check.

**How p-values will be derived (Part C).**
For any test bin $k$, the two-sided empirical p-value is the fraction of null statistic values at least as extreme:
$$p_k = \frac{\bigl|\{j \in \text{ctrl} : |T_j| \geq |T_k|\}\bigr|}{|\text{ctrl}|}$$
Because the null is approximately normal (confirmed below), this is equivalent to $p_k = 2\Phi(-|z_k|)$ where $z_k = T_k / \hat{\sigma}_0$ and $\hat{\sigma}_0$ is the SD of the control bins. Both approaches give consistent results; the empirical version makes no distributional assumption.

```{r neg-ctrl-strategies}
# Strategy A: global MAD-3
outlier_thresh <- 3 * mad(df$cn[nz], constant = 1)
stratA <- nz & abs(df$cn - 1) <= outlier_thresh

# Strategy B: local stability in log space
win_med_log <- roll_stat(log_cn, median)
win_sd_log  <- roll_stat(log_cn, sd)
stratB <- nz & !is.na(win_med_log) & !is.na(win_sd_log) &
          abs(win_med_log) <= 0.095 & win_sd_log <= 0.14

cat(sprintf("Strategy A: %d bins (%.0f%% of non-zero)\n",
            sum(stratA), 100 * sum(stratA) / sum(nz)))
cat(sprintf("Strategy B: %d bins (%.0f%% of non-zero)\n",
            sum(stratB), 100 * sum(stratB) / sum(nz)))
```

```{r null-distributions, fig.width=12, fig.height=8, dev="png", dpi=180}
stat_list  <- list(S1 = df$S1, S2 = df$S2, S3 = df$S3)
stat_labs  <- c("S1: log(a-hat)", "S2: trimmed mean", "S3: Huber")
strat_list <- list(A = stratA, B = stratB)
strat_labs <- c("Strategy A (MAD-3)", "Strategy B (local stability)")
strat_cols <- c("#56B4E9", "#009E73")

par(mfrow = c(2, 3), mar = c(4, 3.5, 2.5, 1))

for (si in 1:2) {
  for (ti in 1:3) {
    mask <- strat_list[[si]] & !is.na(stat_list[[ti]])
    v    <- stat_list[[ti]][mask]
    v    <- pmin(pmax(v, -1), 1.5)
    hist(v, breaks = 80, freq = FALSE,
         col = adjustcolor(strat_cols[si], 0.6), border = NA,
         main = sprintf("%s\n%s", strat_labs[si], stat_labs[ti]),
         xlab = "Statistic under H0", cex.main = 0.78,
         xlim = c(-0.8, 0.8))
    abline(v = 0, col = "red", lwd = 2, lty = 2)
    legend("topright", bty = "n", cex = 0.75,
           legend = sprintf("n=%d\nSD=%.3f\nMAD=%.3f",
                            length(v), sd(v, na.rm=TRUE), mad(v, na.rm=TRUE)))
  }
}
```

```{r qqplots, fig.width=12, fig.height=8, dev="png", dpi=180}
par(mfrow = c(2, 3), mar = c(4, 3.5, 2.5, 1))

for (si in 1:2) {
  for (ti in 1:3) {
    mask <- strat_list[[si]] & !is.na(stat_list[[ti]])
    v    <- stat_list[[ti]][mask]
    qqnorm(v, pch = 20, cex = 0.2, col = adjustcolor(strat_cols[si], 0.3),
           main = sprintf("QQ: %s\n%s", strat_labs[si], stat_labs[ti]),
           cex.main = 0.78)
    qqline(v, col = "red", lwd = 2)
  }
}
```

```{r null-summary-table}
rows <- list()
for (si in 1:2) {
  for (ti in 1:3) {
    mask <- strat_list[[si]] & !is.na(stat_list[[ti]])
    v    <- stat_list[[ti]][mask]
    rows[[length(rows)+1]] <- data.frame(
      Strategy  = names(strat_list)[si],
      Statistic = names(stat_list)[ti],
      n         = length(v),
      Mean      = round(mean(v), 4),
      SD        = round(sd(v),   4),
      MAD       = round(mad(v),  4)
    )
  }
}
summary_tbl <- do.call(rbind, rows)
knitr::kable(summary_tbl, caption = "Null distribution summary: 3 statistics x 2 strategies")
```

**Conclusion.**
Strategy B + S3 (Huber) gives the tightest, most normal null (SD = 0.042, nearly linear QQ plot) and is our chosen combination for parts C onward.
Strategy A + S1 serves as a baseline that makes the benefit of both robustness and stricter control selection clearly visible.

------------------------------------------------------------------------

## External sources statement

We used AI (Claude) to design the negative control selection strategies (exploratory analysis comparing MAD-3 vs local stability filters) and to structure this document.
Statistical choices (Huber k=1.345, trimming fraction, window width) and interpretation are our own.
