---
title: "HW9"
author: "Group 12 - Shir, Omer & Ido"
date: "16/06/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
always_allow_html: true
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

------------------------------------------------------------------------

### Paths and Data

```{r setup, message=FALSE, warning=FALSE}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }
if (!require('MASS'))       { install.packages('MASS');       library('MASS') }

source("../utils.R")
source("../lab5/lab5_utils.R")

healthy_file <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
tumor_file   <- "../lab7/data/TCGA-13-0723-01A_lib2_all_chr1.forward"
chr1_file    <- "../lab1/data/chr1_str.rda"
```

```{r load-data, cache=TRUE}
healthy_reads <- fread(healthy_file, col.names = c("Chr", "Loc", "Length"))
tumor_reads   <- fread(tumor_file,   col.names = c("Chr", "Loc", "Length"))
load(chr1_file)
```

```{r bin-and-correct, cache=TRUE}
BIN <- 2500

add_pos <- function(df, beg) {
  df$pos <- beg + (seq_len(nrow(df)) - 1) * BIN + BIN / 2
  df
}

# Two arms around centromere (120-130 Mb excluded), both samples
df_h <- rbind(
  add_pos(bin_data(healthy_reads, chr1, 5000001L,   119990000L, BIN), 5000001L),
  add_pos(bin_data(healthy_reads, chr1, 130000001L, 242000000L, BIN), 130000001L)
)
df_t <- rbind(
  add_pos(bin_data(tumor_reads, chr1, 5000001L,   119990000L, BIN), 5000001L),
  add_pos(bin_data(tumor_reads, chr1, 130000001L, 242000000L, BIN), 130000001L)
)

# Per-sample GC spline + plug-in estimator (lab 8 method b)
eps <- 1
gc_correct <- function(df) {
  nz   <- df$reads > 0
  fit  <- fit_gc_spline(df[nz, ])
  fhat <- pmax(predict(fit, data.frame(gc = df$gc)), 0)
  raw  <- (df$reads + eps) / (fhat + eps)
  raw / median(raw[nz])
}
df_h$cn <- gc_correct(df_h)
df_t$cn <- gc_correct(df_t)

# Two-sample estimator (lab 8 method d): GC-corrected tumor / GC-corrected healthy
df_t$cn_2s <- df_t$cn / df_h$cn

nz_h <- df_h$reads > 0
nz_t <- df_t$reads > 0
nz   <- nz_h & nz_t   # bins with reads in both samples

# alias so rest of code uses df / df$cn referring to working data frame
df <- df_t

cat(sprintf("Total bins: %d  |  both non-zero: %d\n", nrow(df), sum(nz)))
```

------------------------------------------------------------------------

## Part A -- Test Statistic

We need a statistic $T$ that summarises whether a genomic region has more or fewer reads than expected under copy number 1, while being **resistant to single noisy bins** -- meaning one extreme bin should not be able to drive the entire region to significance on its own.

We use the **two-sample GC-corrected estimator** (Lab 8 method d): $\hat{a}_k = \hat{a}^\text{tumor}_k / \hat{a}^\text{healthy}_k$, where each sample is corrected by its own GC spline and median-normalised.
Dividing cancels shared GC bias and library-specific technical effects, leaving a cleaner copy-number signal than single-sample correction alone.

We then take $\log_2(\hat{a}_k)$: under $H_0$ (CN = 1) this is centred at 0, gains are positive, losses are negative, and the $\log_2$ scale has a natural interpretation (a value of 1 means a doubling; $-1$ means a halving).

We considered three candidates, all computed as a rolling aggregate over a window of $W = 11$ bins ($\approx 27.5$ kb):

| Statistic | Formula | Notes |
|----|----|----|
| **S1**: plain log$_2$ | $\log_2(\hat{a}_k)$ -- per bin, no aggregation | Baseline |
| **S2**: trimmed mean | $\overline{\log_2(\hat{a})}_{10\%}$ over window | Hard-discards top/bottom 10% of bins |
| **S3**: Huber location | $\hat{\mu}_\text{Huber}(\log_2(\hat{a}),\, k=1.345)$ over window | Continuously downweights outlier bins |

We evaluated all three against the empirical null (Part B).
S2 and S3 produce noticeably narrower, more compressed null distributions when combined with the local-stability negative control filter: the filter already removes noisy bins, so a further robust aggregation over-compresses the tails.
S1 -- plain $\log_2(\hat{a}_k)$ -- keeps the most natural spread.
We carry all three forward and defer the final choice to Part B, where a calibration test on held-out control bins selects between them; the "resistance to single noisy bins" requirement is in any case addressed by the negative control selection (Part B) rather than by the statistic alone.

```{r define-statistics}
W    <- 11
HALF <- (W - 1) / 2

log2_cn <- log2(pmax(df$cn_2s, 1e-6))   # log2 of two-sample a-hat

# Rolling window helper (kept for S2/S3 comparison plots)
roll_stat <- function(x, fun) {
  out <- rep(NA_real_, length(x))
  for (i in seq_along(x)) {
    idx <- max(1, i - HALF) : min(length(x), i + HALF)
    sub <- x[idx][nz[idx]]
    if (length(sub) >= 5) out[i] <- fun(sub)
  }
  out
}

df$S1 <- log2_cn
df$S2 <- roll_stat(log2_cn, function(x) mean(x, trim = 0.10))
df$S3 <- roll_stat(log2_cn, function(x) huber(x, k = 1.345)$mu)
```

------------------------------------------------------------------------

## Part B -- Negative Control Regions and Null Distribution

**How we select the negative controls.** We need genomic regions where CN = 1 with high confidence, so that the statistic values computed there serve as a direct sample from $H_0$.
We use the paired healthy tissue sample (TCGA-13-0723-10B): a normal blood cell has CN = 1 almost everywhere on chr1, making the healthy genome our natural candidate pool.

We apply the local stability filter (**Strategy B**) directly on the **two-sample ratio** $\log_2(\hat{a}_k)$: a bin is kept as a negative control only if its 11-bin neighbourhood simultaneously has a median near 0 **and** a low local SD.
We set the two cut-offs in interpretable linear-ratio units and then convert to $\log_2$: the local median CN must stay within $\pm 10\%$ of the no-change baseline ($|\tilde{T}| \leq \log_2 1.10 \approx 0.137$), and the local scatter must stay under $\pm 15\%$ (SD $\leq \log_2 1.15 \approx 0.202$).
Filtering on the ratio rather than the healthy sample alone is the correct choice: $H_0$ is "no tumor-specific CN change" (ratio = 1), not "healthy CN = 1 absolutely." This also handles germline CNVs correctly — if both samples carry a duplication at some locus, the ratio is still 1 and the bin is a valid negative control.
Bins at tumor CN-event boundaries are excluded because their neighbours reveal the disruption even if the bin itself lands near 1.
Approximately 37% of non-zero bins pass this filter, giving \~31,000 control bins across both chromosome arms.

**How we model** $H_0$.
The $\log_2(\hat{a}_k)$ values of the selected control bins are a direct sample from the distribution of the statistic when CN = 1 -- we use them directly as the empirical null, without assuming any parametric form.
The histograms and QQ plots below characterise this distribution for each of the three candidate statistics.

**How p-values are derived (Part C).** For any test bin $k$ with statistic value $T_k$, the two-sided empirical p-value is the fraction of null values at least as extreme: $$p_k = \frac{\bigl|\{j \in \text{ctrl} : |T_j| \geq |T_k|\}\bigr|}{|\text{ctrl}|}$$ With \$\sim\$31,000 control bins this resolves p-values down to $\approx 3 \times 10^{-5}$, sufficient for testing at $\alpha = 0.01$ after multiple testing correction.
No normality assumption is required.

```{r neg-ctrl}
# Stability filter runs on the two-sample ratio (log2_cn = log2(tumor/healthy)).
# H0 is "no tumor-specific CN change" (ratio = 1), so the filter should assess
# whether the ratio is locally stable near 0 -- not whether the healthy alone is.
# This also correctly handles germline CNVs: if both samples have CN=2, ratio=1
# and the bin is still a valid negative control.
# Thresholds set in interpretable linear-ratio units, then converted to log2:
#   median within +/-10% of the no-change baseline; local scatter under +/-15%.
med_thresh <- log2(1.10)   # ~0.137
sd_thresh  <- log2(1.15)   # ~0.202

win_med_log <- roll_stat(log2_cn, median)
win_sd_log  <- roll_stat(log2_cn, sd)
stratB <- nz & !is.na(win_med_log) & !is.na(win_sd_log) &
          abs(win_med_log) <= med_thresh & win_sd_log <= sd_thresh

cat(sprintf("Strategy B (local stability): %d bins (%.0f%% of non-zero)\n",
            sum(stratB), 100 * sum(stratB) / sum(nz)))
```

```{r negctrl-scatter, fig.width=9, fig.height=4, dev="png", dpi=180}
# Chromosome-wide scatter: control bins highlighted, excluded in grey
set.seed(42)
idx_all <- which(nz)
s       <- sample(idx_all, min(25000, length(idx_all)))
x_mb    <- df$pos[s] / 1e6
y_cn    <- pmin(log2_cn[s], 3)

col_vec <- ifelse(stratB[s],
                  adjustcolor("#009E73", 0.5),
                  adjustcolor("#999999", 0.2))

par(mar = c(4.5, 4, 2.5, 1), mgp = c(2.2, 0.7, 0))
plot(x_mb, y_cn, pch = 20, cex = 0.25, col = col_vec,
     xlab = "Position (Mb)", ylab = expression(log[2](hat(a)[k])),
     main = "Negative control bins across chr1  (green = control, grey = excluded)",
     ylim = c(-2, 3))
abline(h = 0, col = "#E69F00", lwd = 1.5, lty = 2)
legend("topright", bty = "n", cex = 0.85,
       legend = c(sprintf("Control (%d bins, %.0f%%)",
                          sum(stratB), 100 * sum(stratB) / sum(nz)), "Excluded"),
       col    = c(adjustcolor("#009E73", 0.8), adjustcolor("#999999", 0.6)),
       pch    = 20, pt.cex = 1.2)
```

```{r null-dist-plot, fig.width=13, fig.height=7.3, dev="png", dpi=180}
stat_list_b <- list(S1 = df$S1, S2 = df$S2, S3 = df$S3)
stat_labs_b <- c("S1: log2(a-hat)", "S2: trimmed mean", "S3: Huber")
cols_b      <- c("#555555", "#56B4E9", "#009E73")

par(mfrow = c(2, 3), mar = c(4.5, 4, 3, 1), mgp = c(2.4, 0.8, 0),
    cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)

for (ti in 1:3) {
  mask  <- stratB & !is.na(stat_list_b[[ti]])
  v     <- stat_list_b[[ti]][mask]
  # xlim: 5 SDs each side so shape is always visible, floor at ±0.5
  hw    <- max(5 * sd(v), 0.5)
  xlim  <- c(-hw, hw)
  xseq  <- seq(xlim[1], xlim[2], length.out = 300)
  h     <- hist(v, breaks = 80, plot = FALSE)
  plot(h, freq = FALSE, col = adjustcolor(cols_b[ti], 0.6), border = NA,
       main = sprintf("Null: Strategy B, %s", stat_labs_b[ti]),
       xlab = "Statistic under H0", xlim = xlim)
  abline(v = 0, col = "red", lwd = 1.5, lty = 2)
  lines(xseq, dnorm(xseq, mean(v), sd(v)), col = "black", lwd = 2)
  legend("topright", bty = "n", cex = 1.0,
         legend = sprintf("n=%d\nSD=%.4f\nMAD=%.4f", length(v), sd(v), mad(v)))
}

for (ti in 1:3) {
  mask <- stratB & !is.na(stat_list_b[[ti]])
  v    <- stat_list_b[[ti]][mask]
  qqnorm(v, pch = 20, cex = 0.3, col = adjustcolor(cols_b[ti], 0.4),
         main = sprintf("QQ: Strategy B, %s", stat_labs_b[ti]))
  qqline(v, col = "red", lwd = 2)
}
```

The histograms are shown with a fitted normal overlay (black line) for reference, and the QQ plots assess normality directly.
S2 and S3 are visibly narrower and more compressed than S1, a consequence of applying robust aggregation on top of bins already filtered for stability.
But comparing shapes to a normal curve is only superficial -- what actually matters is whether each statistic produces **calibrated** p-values.
We test this directly below.

### Validating calibration (split-half)

A statistic is only usable if, on data that truly follows $H_0$, its p-values are **Uniform(0,1)** -- equivalently, the realised false-positive rate at level $\alpha$ equals $\alpha$.
We can check this with the control bins alone, no labelled events required:

1.  Split the control bins into **interleaved genomic blocks** (block size 100 bins $\gg$ window $W=11$, so the two sets never share a window). Even blocks build the empirical null; odd blocks are held out as "test" bins.
2.  The held-out bins are themselves CN = 1, so under a correctly calibrated test their two-sided empirical p-values must be uniform.
3.  We measure the deviation from uniform (Kolmogorov--Smirnov $D$) and the realised false-positive rate at $\alpha = 0.01$ and $0.05$.

A statistic whose null is artificially narrow (S2/S3) will make ordinary CN = 1 fluctuations look significant, inflating the false-positive rate above $\alpha$ -- which disqualifies it regardless of how its histogram looks.

```{r calibration}
# Interleaved-block split so null-building and test bins never share a window
ctrl_idx <- which(stratB)
blk       <- 100                          # block size (bins) >> W = 11
block_id  <- ctrl_idx %/% blk
null_ix   <- ctrl_idx[block_id %% 2 == 0] # even blocks -> build null
test_ix   <- ctrl_idx[block_id %% 2 == 1] # odd blocks  -> held-out CN=1 tests

# Add-one two-sided empirical p-value (never zero, valid by construction)
emp_p2 <- function(t_test, t_null) {
  n <- length(t_null)
  at <- abs(t_test)
  vapply(at, function(a) (1 + sum(abs(t_null) >= a)) / (n + 1), numeric(1))
}

p_lists <- list()
calib_rows <- list()
for (ti in seq_along(stat_list_b)) {
  v_null <- stat_list_b[[ti]][null_ix]; v_null <- v_null[is.finite(v_null)]
  v_test <- stat_list_b[[ti]][test_ix]; v_test <- v_test[is.finite(v_test)]
  m <- median(v_null)                     # H0 centre, estimated from the null
  p <- emp_p2(v_test - m, v_null - m)
  p_lists[[ti]] <- p
  calib_rows[[ti]] <- data.frame(
    Statistic = names(stat_list_b)[ti],
    n_test    = length(p),
    FPR_0.01  = round(mean(p < 0.01), 4),
    FPR_0.05  = round(mean(p < 0.05), 4),
    KS_D      = round(suppressWarnings(ks.test(p, "punif")$statistic), 4)
  )
}
calib_tbl <- do.call(rbind, calib_rows)
rownames(calib_tbl) <- NULL
knitr::kable(calib_tbl,
             col.names = c("Statistic", "n (test)", "FPR @ 0.01", "FPR @ 0.05", "KS D"),
             caption = "Calibration on held-out CN=1 bins. Well-calibrated: FPR approx alpha, KS D small.")
```

```{r calibration-plot, fig.width=12, fig.height=5, dev="png", dpi=180}
cols_b <- c("#555555", "#56B4E9", "#009E73")
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1), mgp = c(2.6, 0.8, 0),
    cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.1)

# Full P-P plot: empirical CDF of p-values vs uniform diagonal
plot(NA, xlim = c(0, 1), ylim = c(0, 1),
     xlab = "Nominal p", ylab = "Observed fraction <= p",
     main = "Calibration (P-P plot): full range")
abline(0, 1, col = "grey50", lwd = 2, lty = 2)
for (ti in 1:3) {
  ps <- sort(p_lists[[ti]])
  lines(ps, (seq_along(ps)) / length(ps), col = cols_b[ti], lwd = 2.5)
}
legend("bottomright", bty = "n", cex = 1.1, lwd = 2.5, col = cols_b,
       legend = names(stat_list_b))

# Zoom into the testing-relevant small-p region
plot(NA, xlim = c(0, 0.1), ylim = c(0, 0.1),
     xlab = "Nominal p", ylab = "Observed fraction <= p",
     main = "Calibration: small-p region")
abline(0, 1, col = "grey50", lwd = 2, lty = 2)
abline(v = 0.01, col = "red", lwd = 1.2, lty = 3)
for (ti in 1:3) {
  ps <- sort(p_lists[[ti]])
  lines(ps, (seq_along(ps)) / length(ps), col = cols_b[ti], lwd = 2.5)
}
legend("topleft", bty = "n", cex = 1.1, lwd = 2.5, col = cols_b,
       legend = names(stat_list_b))
```

**Result.** All three statistics are calibrated: every realised false-positive rate sits at or below its nominal $\alpha$, and the P--P curves track the diagonal.
This is expected -- empirical p-values are self-calibrating by construction, because each statistic is ranked against *its own* null, so the over-compression of S2/S3 cancels between the null and the test bins.
The narrow histograms of S2/S3 therefore do **not** cause false positives; superficial non-normality was a red herring.

The check still discriminates, though.
**S1 is the best calibrated** (KS $D \approx 0.009$, with FPR almost exactly $0.01$ and $0.05$), tracking the uniform diagonal nearly perfectly.
S2 and S3 are slightly **conservative** -- their realised FPR falls *below* $\alpha$ and their P--P curves bow under the diagonal -- because the heavy aggregation discretises the statistic and inflates ties.
Conservatism is safe but costs power: at a fixed $\alpha$, a conservative test detects fewer true events.
Calibration alone does not strictly eliminate any candidate, but it gives us a principled basis to choose: **S1 is both the best-calibrated (tightest to the uniform diagonal) and the simplest** -- a single-bin statistic with no aggregation -- while the conservatism of S2/S3 implies they would detect fewer true events at the same $\alpha$.
We therefore carry **S1** forward as our statistic for the genome-wide scan in Part C.

------------------------------------------------------------------------

## Part C - Computing P-values

Based on our conclusion in Part B, we proceed with **Statistic S1** (plain log) and **Strategy B** (local stability filter).
We use the control bins identified by Strategy B to define our empirical null distribution.
For every testable bin on the chromosome -- one with reads in both the tumor and the matched healthy sample -- we calculate a two-sided empirical p-value.
The empirical null is summarised below: it is built from the Strategy-B control bins, is centred essentially at 0 (confirming the $\log_2$ ratio is unbiased under $H_0$), and is narrow enough to resolve very small p-values.

```{r compute-pvalues}
# Isolate the empirical null distribution values
null_stats <- df$S1[stratB & !is.na(df$S1)]
n_null     <- length(null_stats)
null_mean  <- mean(null_stats)
null_sd    <- sd(null_stats)

# Report the null summary as a clean table
null_summary <- data.frame(
  Quantity = c("Control bins (null size)", "Null mean (S1)", "Null SD (S1)",
               "Smallest attainable p-value"),
  Value = c(format(n_null, big.mark = ","),
            sprintf("%.5f", null_mean),
            sprintf("%.5f", null_sd),
            signif(1 / n_null, 2))
)
knitr::kable(null_summary, align = c("l", "r"),
             caption = "Empirical null distribution of S1 on the control bins.")

# Calculate empirical two-sided p-values.
# p_k = |{j in ctrl : |T_j| >= |T_k|}| / |ctrl|
# IMPORTANT: only test bins with reads in *both* samples (nz). A bin with zero
# coverage in a sample gets an eps-smoothed ratio far from 1 and would read as a
# huge "loss" -- e.g. the unmappable pericentromeric band -- so we leave such
# bins untested (p = NA) rather than calling them significant.
abs_null <- abs(null_stats)

df$p_val <- NA_real_
df$p_val[nz] <- sapply(abs(df$S1[nz]), function(x) {
  if (is.na(x)) return(NA_real_)
  sum(abs_null >= x) / n_null
})

# Prevent p=0 for FDR correction stability
min_p <- 1 / n_null
df$p_val[nz] <- pmax(df$p_val[nz], min_p)

# Preview the most extreme bins (smallest p-values), as a clean table
top_bins <- df[order(df$p_val), ][1:10, ]
top_tbl <- data.frame(
  `Position (Mb)` = round(top_bins$pos / 1e6, 3),
  `Tumor reads`   = top_bins$reads,
  `Ratio a-hat`   = round(top_bins$cn_2s, 3),
  `S1 = log2`     = round(top_bins$S1, 3),
  `p-value`       = signif(top_bins$p_val, 2),
  check.names = FALSE
)
knitr::kable(top_tbl, align = "r",
             caption = "Ten most extreme bins by empirical p-value (smallest p first).")
```

### How different is the parametric alternative?

We could instead assume the statistic is Normal under $H_0$ and read p-values off the fitted curve: with $\mu_0, \sigma_0$ estimated from the same control bins, the two-sided parametric p-value is $p_k = 2\,\bigl[1 - \Phi(|T_k - \mu_0| / \sigma_0)\bigr]$.
It is natural to ask how much this choice actually matters -- do the two approaches flag different bins?

```{r param-pvalues}
# Parametric (Normal) p-values from the same control-bin mean/SD, for comparison.
z_score        <- (df$S1 - null_mean) / null_sd
df$p_val_param <- NA_real_
df$p_val_param[nz] <- 2 * pnorm(-abs(z_score[nz]))

# BH on both, then compare the significant calls
sig_emp   <- !is.na(df$p_val)       & p.adjust(df$p_val,       method = "BH") < 0.01
sig_param <- !is.na(df$p_val_param) & p.adjust(df$p_val_param, method = "BH") < 0.01

cmp_tbl <- data.frame(
  Quantity = c("Nominally significant (p < 0.01)",
               "Significant after BH (FDR < 0.01)",
               "Called by both",
               "Called by empirical only",
               "Called by parametric only"),
  Count = c(sprintf("%d  /  %d", sum(df$p_val < 0.01, na.rm = TRUE),
                                  sum(df$p_val_param < 0.01, na.rm = TRUE)),
            sprintf("%d  /  %d", sum(sig_emp), sum(sig_param)),
            sum(sig_emp & sig_param),
            sum(sig_emp & !sig_param),
            sum(sig_param & !sig_emp))
)
knitr::kable(cmp_tbl, align = c("l", "r"),
             col.names = c("Quantity", "Empirical / Parametric"),
             caption = "Bins flagged by the empirical vs parametric (Normal) p-values.")
```

```{r param-compare-plot, fig.width=8, fig.height=4, dev="png", dpi=180}
# Bins sorted by empirical p-value vs rank, with the BH significance cutoffs marked.
# Colour carries the empirical-vs-parametric difference: which significant bins
# both methods agree on, vs the ones only the empirical test keeps.
alpha <- 0.01
pv <- df$p_val[nz]; se <- sig_emp[nz]; sp <- sig_param[nz]
m  <- length(pv)
ord <- order(pv)
pv_s <- pv[ord]; se_s <- se[ord]; sp_s <- sp[ord]
rank <- seq_len(m)

col_cls <- ifelse(se_s & sp_s, "both",
           ifelse(se_s & !sp_s, "emp_only", "none"))
pal <- c(both = "#1b9e77", emp_only = "#e31a1c", none = adjustcolor("grey55", 0.5))

nsig <- sum(se); nsig_par <- sum(sp)
xmax <- nsig + 400; sel <- rank <= xmax       # zoom to the decision region

par(mar = c(4.5, 4.7, 3, 1), mgp = c(2.7, 0.8, 0))
plot(rank[sel], pv_s[sel], pch = 20, cex = 0.5, col = pal[col_cls[sel]],
     xlab = "Bin rank (sorted by empirical p-value)", ylab = "Empirical p-value",
     main = "BH significance cutoff on empirical p-values", ylim = c(0, max(pv_s[sel])))
abline(v = nsig_par, col = "#999999", lty = 3, lwd = 1.3)
abline(v = nsig,     col = "#7570b3", lty = 2, lwd = 1.6)
text(nsig, max(pv_s[sel]) * 0.92, sprintf("empirical BH cutoff\n(%d bins)", nsig),
     col = "#7570b3", adj = c(1.05, 1), cex = 0.8)
text(nsig_par, max(pv_s[sel]) * 0.40, sprintf("parametric\ncutoff (%d)", nsig_par),
     col = "#666666", adj = c(1.05, 1), cex = 0.78)
legend("topleft", bty = "n", pch = 20, pt.cex = 1.3, cex = 0.9,
       col = c("#1b9e77", "#e31a1c", "grey55"),
       legend = c("Sig in both methods", "Sig empirical-only (573)", "Not significant"))
```

The BH procedure keeps every bin up to the largest rank whose sorted p-value still satisfies the FDR criterion; the dashed line marks that empirical cutoff (2,913 bins) and the dotted line the parametric one (2,340).
**For the vast majority of bins the choice of p-value does not matter** -- the two methods agree on the 2,340 calls with the smallest p-values (green), and on the tens of thousands of clearly non-significant bins (grey).
The disagreement is small and entirely one-sided: the empirical test additionally keeps 573 bins (red), while the parametric keeps **none** that the empirical misses, so the parametric calls are a strict subset.
Those 573 red bins are exactly the marginal ones between the two cutoffs: the more conservative Normal model drops them, costing power, without any compensating gain (in the far tail it instead becomes wildly over-confident, assigning the single most extreme bin $p \approx 10^{-133}$ -- a precision \$\sim\$31,000 control observations cannot support, whereas the empirical p-value honestly bottoms out at $1/m \approx 3.2\times10^{-5}$).
Since the empirical null also needs no distributional assumption, we keep it for the genome-wide scan.

------------------------------------------------------------------------

## Part D - FDR Correction and Segmentation

At the nominal level $\alpha = 0.01$, before any correction, a large number of bins are flagged (the uncorrected count is reported below, answering part ד) -- but with tens of thousands of bins tested, roughly $\alpha \times N$ of those are expected false positives purely by chance, so the raw count is not trustworthy.

**The property the correction preserves (part ה).** We use the **Benjamini--Hochberg (BH)** procedure, which controls the **False Discovery Rate (FDR)** -- the *expected proportion of false positives among the bins we call significant*.
Setting the BH target to $0.01$ guarantees that, on average, at most $1\%$ of our reported significant bins are false discoveries.
This is deliberately different from Bonferroni, which controls the **family-wise error rate** (the probability of *even one* false positive anywhere) and would be far too strict here: with \$\sim\$85,000 tests Bonferroni's per-test threshold collapses to $\approx 10^{-7}$ and would discard almost all of the true, weaker copy-number events.
Because we are screening a whole chromosome for *many* real alterations rather than guarding against a single false call, FDR control is the appropriate trade-off -- it tolerates a small, known fraction of false positives in exchange for much greater power to detect true events.
Adjacent significant windows are then grouped into continuous genomic segments (events).

```{r fdr-correction}
alpha <- 0.01

# Apply Benjamini-Hochberg FDR correction
df$p_adj <- p.adjust(df$p_val, method = "BH")
df$is_sig <- !is.na(df$p_adj) & df$p_adj < alpha

# Determine the type of change (Gain vs Loss)
# Since S1 = log2(a_hat), positive is gain, negative is loss.
df$change_type <- "Normal"
df$change_type[df$is_sig & df$S1 > 0] <- "Gain"
df$change_type[df$is_sig & df$S1 < 0] <- "Loss"

# Headline counts: tested bins, nominal hits (part D), and post-correction calls
n_tested  <- sum(!is.na(df$p_val))
n_raw_sig <- sum(df$p_val < alpha, na.rm = TRUE)   # before correction (spec part D)
n_sig     <- sum(df$is_sig, na.rm = TRUE)
count_tbl <- data.frame(
  Quantity = c("Testable bins (reads in both samples)",
               sprintf("Nominally significant (p < %.2f, uncorrected)", alpha),
               sprintf("Significant after BH (FDR < %.2f)", alpha),
               "  - of which Gains (ratio > 1)",
               "  - of which Losses (ratio < 1)"),
  Count = c(n_tested, n_raw_sig, n_sig,
            sum(df$change_type == "Gain"), sum(df$change_type == "Loss"))
)
knitr::kable(count_tbl, align = c("l", "r"),
             caption = "Significant-bin counts before and after multiple-testing correction.")

# Group adjacent significant bins into segments using rle
rle_res <- rle(df$change_type)
end_idx <- cumsum(rle_res$lengths)
start_idx <- c(1, head(end_idx, -1) + 1)

segments <- data.table(
  change_type = rle_res$values,
  start_idx = start_idx,
  end_idx = end_idx,
  n_bins = rle_res$lengths
)

# Filter for significant segments only
sig_segments <- segments[change_type != "Normal"]

# Calculate genomic coordinates and summary stats for each segment.
# Report the two-sample ratio cn_2s (= tumor/healthy) -- that is the quantity the
# whole test is built on. The single-sample tumor cn ignores the healthy baseline
# and can disagree in sign with the event call.
sig_segments[, `:=`(
  start_mb = df$pos[start_idx] / 1e6,
  end_mb   = df$pos[end_idx] / 1e6,
  size_kb  = (df$pos[end_idx] - df$pos[start_idx] + BIN) / 1e3,
  mean_ratio = mapply(function(s, e) mean(df$cn_2s[s:e], na.rm = TRUE), start_idx, end_idx),
  min_padj = mapply(function(s, e) min(df$p_adj[s:e], na.rm = TRUE), start_idx, end_idx)
)]

# Clean up table for display, filtering for segments > 10kb to remove single-bin noise
segments_display <- sig_segments[size_kb > 10, .(
  Event = change_type,
  Start_Mb = round(start_mb, 2),
  End_Mb = round(end_mb, 2),
  Size_kb = round(size_kb, 1),
  Bins = n_bins,
  Mean_ratio = round(mean_ratio, 3),
  Min_FDR = signif(min_padj, 3)
)]

knitr::kable(segments_display, caption = "Significant Copy Number Alteration Segments (FDR < 0.01, Size > 10kb)")
```

## Part E - Visualization

To visualize the results of our copy number variation analysis, we plot the `S1` statistic ($\log(\hat{a}_k)$) across the chromosome, highlighting the regions identified as statistically significant amplifications (Gain, CN \> 1) and deletions (Loss, CN \< 1).

We provide two views: a static genome-wide plot (below) that renders in any output format, and an interactive Plotly version (rendered only in the HTML build) that supports zoom, pan and per-bin hover -- the "movable graph" the task allows as an alternative to a region list.

```{r cnv-static-plot, fig.width=12, fig.height=4.5, dev="png", dpi=180}
# Static genome-wide view -- always renders (PDF and HTML).
set.seed(42)
norm_idx <- which(df$change_type == "Normal" & !is.na(df$S1))
norm_s   <- sample(norm_idx, min(15000, length(norm_idx)))
gain_idx <- which(df$change_type == "Gain")
loss_idx <- which(df$change_type == "Loss")

par(mar = c(4.2, 4, 2.5, 1), mgp = c(2.3, 0.7, 0))
plot(df$pos[norm_s] / 1e6, pmin(pmax(df$S1[norm_s], -3), 3),
     pch = 20, cex = 0.25, col = adjustcolor("grey60", 0.35),
     xlab = "Genomic position (Mb)", ylab = expression(S1 == log[2](hat(a)[k])),
     main = "Chr1: detected copy-number variants (FDR < 0.01)",
     ylim = c(-3, 3), xlim = range(df$pos / 1e6))
abline(h = 0, col = "grey30", lty = 2)
points(df$pos[loss_idx] / 1e6, pmin(pmax(df$S1[loss_idx], -3), 3),
       pch = 20, cex = 0.5, col = adjustcolor("#1f78b4", 0.7))
points(df$pos[gain_idx] / 1e6, pmin(pmax(df$S1[gain_idx], -3), 3),
       pch = 20, cex = 0.5, col = adjustcolor("#e31a1c", 0.7))
legend("topright", bty = "n", pch = 20, pt.cex = 1.2,
       col = c("grey60", "#e31a1c", "#1f78b4"),
       legend = c("Normal", "Gain (ratio > 1)", "Loss (ratio < 1)"))
```

It is also worth seeing the **raw tumor read counts** directly, with each bin coloured by its anomaly call.
This shows the detection acting on the actual coverage: significant losses sit in the low read-count bins and gains in the high ones, while the grey background stays near the chromosome-wide median (\~216 reads / 2.5 kb bin).

```{r cnv-reads-plot, fig.width=12, fig.height=4.5, dev="png", dpi=180}
# Raw tumor read counts per bin, coloured by CNV call. Pure scatter: no lines,
# so the centromere gap stays empty (no interpolation across it).
ycap <- quantile(df$reads[nz], 0.995)        # cap y so a few huge bins don't dominate
x_mb <- df$pos / 1e6
y    <- pmin(df$reads, ycap)

# Draw Normal first, then anomalies on top so the calls are not buried
ord  <- order(df$change_type != "Normal")
cols <- c(Normal = adjustcolor("grey65", 0.35),
          Gain   = adjustcolor("#e31a1c", 0.8),
          Loss   = adjustcolor("#1f78b4", 0.8))

par(mar = c(4.2, 4, 2.5, 1), mgp = c(2.3, 0.7, 0))
plot(x_mb[ord], y[ord], pch = 20,
     cex = ifelse(df$change_type[ord] == "Normal", 0.2, 0.5),
     col = cols[df$change_type[ord]],
     xlab = "Genomic position (Mb)", ylab = "Tumor reads per 2.5 kb bin",
     main = "Chr1: tumor read counts, coloured by CNV call (FDR < 0.01)",
     ylim = c(0, ycap), xlim = range(x_mb))
legend("topright", bty = "n", pch = 20, pt.cex = 1.4,
       col = cols[c("Normal", "Gain", "Loss")],
       legend = c("Normal", "Gain (ratio > 1)", "Loss (ratio < 1)"))
```

The interactive version below downsamples the "Normal" bins (keeping all significant ones) so the widget stays responsive.
It is always exported to a self-contained `AnomalyDetections.html` file (which can be hosted and linked from a slide deck), and is shown inline only in the HTML build.

```{r interactive-plot, warning=FALSE, message=FALSE}
if (!require('plotly'))      { install.packages('plotly');      library('plotly') }
if (!require('htmlwidgets')) { install.packages('htmlwidgets'); library('htmlwidgets') }

# Downsample 'Normal' bins to keep the Plotly interactive plot responsive
set.seed(42)
normal_bins <- which(df$change_type == "Normal" & !is.na(df$S1))
sampled_normal <- sample(normal_bins, min(10000, length(normal_bins)))
sig_bins <- which(df$change_type != "Normal" & !is.na(df$S1))

# Combine sampled normal bins and all significant bins
plot_df <- df[c(sampled_normal, sig_bins), ]
# Reorder by position so the hover behaves nicely
plot_df <- plot_df[order(plot_df$pos), ]

fig <- plot_ly(
  data = plot_df,
  x = ~pos / 1e6,
  y = ~S1,
  type = 'scatter',
  mode = 'markers',
  color = ~change_type,
  colors = c("Normal" = "grey", "Gain" = "red", "Loss" = "blue"),
  marker = list(size = 3, opacity = 0.5),
  text = ~paste("Position:", round(pos / 1e6, 3), "Mb<br>S1 Stat:", round(S1, 4), "<br>Status:", change_type),
  hoverinfo = "text"
) %>%
  layout(
    title = "Chromosome 1: Detected Copy Number Variations",
    xaxis = list(title = "Genomic Position (Mb)"),
    yaxis = list(title = "S1 Statistic (log(a-hat))"),
    legend = list(title = list(text = "<b> Status </b>"))
  )

# Always write the standalone interactive file (for hosting / slide linking)
htmlwidgets::saveWidget(fig, "AnomalyDetections.html", selfcontained = TRUE)

# Show inline only in HTML output; the PDF uses the static plot above
if (knitr::is_html_output()) fig else knitr::asis_output("")
```

These plots bring together our robust statistic mapping, the empirical null calibration, and the Benjamini-Hochberg FDR filtering across the entirety of Chromosome 1.

Several features are visible across the chromosome:

-   **The Centromere Gap:** The clear window between roughly 120 Mb and 130 Mb is the centromeric region, excluded during binning because of known reference-mapping assembly gaps. We only test bins with reads in *both* samples, so the unmappable pericentromeric flanks are left untested rather than mis-called as losses.
-   **The Background Envelope:** The dense band of grey markers around $S1 = 0$ is the copy-number-neutral background ($H_0$). Because the empirical null is built from the locally stable control bins (Strategy B), the bulk of both arms sits flat near zero, as expected.
-   **Significant Losses (Blue):** The clearest focal deletion is near 28.5 Mb (mean ratio $\approx 0.54$, a roughly two-fold loss), and a cluster of significant losses appears around 141--142 Mb. These are evaluated on the two-sample ratio, so they reflect tumor-specific change relative to the matched healthy sample, not absolute tumor coverage.
-   **Significant Gains (Red):** Bins rising above the baseline envelope are candidate tumor-specific amplifications; the calls are more scattered and focal than the losses.

Because we test the single-bin statistic $S1$, individual calls are at fine resolution; the `rle` segmentation then groups runs of same-direction significant bins into the events listed in the table above.
The size filter ($>10$ kb) in that table suppresses isolated single-bin calls, while the plots show every significant bin.

------------------------------------------------------------------------

## Part F -- Local Zoom-In Plots

We zoom into four candidate areas (75 Mb, 100 Mb, 145 Mb, 190 Mb), comparing raw tumor read counts (coloured by CNV call) against the matched healthy sample.
A 51-bin rolling median ($\approx$ 127.5 kb) is overlaid for each sample to reveal regional trends through bin-level noise.

```{r zoom-setup, message=FALSE, warning=FALSE}
if (!require('ggplot2'))   { install.packages('ggplot2');   library('ggplot2') }
if (!require('dplyr'))     { install.packages('dplyr');     library('dplyr') }
if (!require('patchwork')) { install.packages('patchwork'); library('patchwork') }
if (!require('zoo'))       { install.packages('zoo');       library('zoo') }

cnv_zoom_df <- data.frame(
  pos_mb        = df$pos / 1e6,
  tumor_reads   = df$reads,
  healthy_reads = df_h$reads,
  S1             = df$S1,
  status         = factor(df$change_type, levels = c("Normal", "Gain", "Loss"))
)

k_roll <- 51
cnv_zoom_df <- cnv_zoom_df %>%
  mutate(
    tumor_rollmed   = rollmedian(tumor_reads,   k = k_roll, fill = NA, align = "center"),
    healthy_rollmed = rollmedian(healthy_reads, k = k_roll, fill = NA, align = "center"),
    S1_rollmed      = rollmedian(S1,            k = k_roll, fill = NA, align = "center")
  )
```

```{r zoom-four-reads, fig.width=12, fig.height=7, dev="png", dpi=180}
plot_reads_zoom <- function(zdf, center_mb, window_mb = 8, title = NULL) {
  if (is.null(title)) title <- paste0("Around ", center_mb, " Mb")
  region <- zdf %>%
    filter(pos_mb >= center_mb - window_mb / 2,
           pos_mb <= center_mb + window_mb / 2)
  pal <- c(Normal = "grey65", Gain = "red", Loss = "blue")
  ggplot(region, aes(x = pos_mb)) +
    geom_point(aes(y = healthy_reads), color = "grey80", alpha = 0.30, size = 0.5) +
    geom_point(aes(y = tumor_reads, color = status), alpha = 0.45, size = 0.6) +
    geom_line(aes(y = tumor_rollmed),   color = "black",   linewidth = 0.8) +
    geom_line(aes(y = healthy_rollmed), color = "#E69F00", linewidth = 0.8, linetype = "dashed") +
    scale_color_manual(values = pal) +
    coord_cartesian(ylim = c(0, 800)) +
    labs(
      title = title,
      x     = "Position (Mb)",
      y     = "Reads per 2.5 kb bin",
      color = "CNV call"
    ) +
    theme_bw()
}

r75  <- plot_reads_zoom(cnv_zoom_df, 75,  8, "Around 75 Mb")
r100 <- plot_reads_zoom(cnv_zoom_df, 100, 8, "Around 100 Mb")
r145 <- plot_reads_zoom(cnv_zoom_df, 145, 8, "Around 145 Mb")
r190 <- plot_reads_zoom(cnv_zoom_df, 190, 8, "Around 190 Mb")

(r75 | r100) / (r145 | r190) +
  plot_annotation(subtitle = "Black line = tumor rolling median  |  Orange dashed = healthy rolling median  |  Points coloured by CNV call")
```

------------------------------------------------------------------------

## External sources statement

We used AI (Claude) to design the negative control selection strategy (exploratory analysis comparing MAD-3 vs local stability filters) and to structure this document.
Statistical choices (window width, stability thresholds) and interpretation are our own.
