---
title: "HW9"
author: "Group 12 - Shir, Omer & Ido"
date: "16/06/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il


```{r cnv-detection-function}
library(data.table)

# Main pipeline to detect copy number variations across the chromosome
find_cnv_regions <- function(data_df, signal_vector, ctrl_start, ctrl_end, alpha = 0.01) {
  
  # Set up a clean working table so we don't mess up our original data
  working_dt <- data.table(chrom_pos = data_df$pos, cn_ratio = signal_vector)
  
  # Part B: Estimate baseline parameters from the negative control region
  is_control <- working_dt$chrom_pos >= ctrl_start & working_dt$chrom_pos <= ctrl_end
  control_values <- working_dt$cn_ratio[is_control]
  control_values <- control_values[is.finite(control_values)] 
  
  baseline_mean <- mean(control_values)
  baseline_sd   <- sd(control_values)
  
  # Part C: Calculate standard Z-scores and two-sided p-values 
  z_scores <- (working_dt$cn_ratio - baseline_mean) / baseline_sd
  working_dt[, p_val := 2 * pnorm(-abs(z_scores))]
  
  # Part D: Track raw significance before any adjustments
  working_dt[, is_sig_raw := p_val < alpha]
  raw_significant_bins <- sum(working_dt$is_sig_raw, na.rm = TRUE)
  
  # Parts E, F: Control the FDR using the Benjamini-Hochberg correction 
  working_dt[, p_adj := p.adjust(p_val, method = "BH")]
  working_dt[, is_sig_adj := p_adj < alpha]
  adj_significant_bins <- sum(working_dt$is_sig_adj, na.rm = TRUE)
  
  # Determine if the change is a deletion or amplification relative to baseline
  working_dt[, change_type := ifelse(cn_ratio > baseline_mean, "Gain (CN > 1)", "Loss (CN < 1)")]
  
  # Part G: Group adjacent significant windows into single continuous events 
  working_dt[, event_id := rleid(is_sig_adj, change_type)]
  
  significant_segments <- working_dt[is_sig_adj == TRUE, .(
    start_mb  = min(chrom_pos) / 1e6,
    end_mb    = max(chrom_pos) / 1e6,
    size_kb   = (max(chrom_pos) - min(chrom_pos)) / 1e3,
    mean_ratio = mean(cn_ratio),
    top_p_adj  = min(p_adj),
    event      = unique(change_type)
  ), by = event_id]
  
  significant_segments[, event_id := NULL] 
  
  # Build the interactive map if plotly is available
  interactive_map <- NULL
  if (requireNamespace("plotly", quietly = TRUE)) {
    library(plotly)
    
    working_dt[, map_color := "Normal"]
    working_dt[is_sig_adj == TRUE & cn_ratio > baseline_mean, map_color := "Gain"]
    working_dt[is_sig_adj == TRUE & cn_ratio <= baseline_mean, map_color := "Loss"]
    
    interactive_map <- plot_ly(
      data = working_dt, x = ~chrom_pos / 1e6, y = ~cn_ratio, 
      type = 'scatter', mode = 'markers', color = ~map_color, 
      colors = c("Normal" = rgb(0,0,0,0.15), "Gain" = "red", "Loss" = "blue"),
      marker = list(size = 4),
      text = ~sprintf("Position: %.3f Mb\nRatio: %.3f\np-adj: %.3e", chrom_pos/1e6, cn_ratio, p_adj)
    ) %>%
    layout(
      title = "Chromosomal Copy Number Alteration Profiles",
      xaxis = list(title = "Genomic Position (Mb)"),
      yaxis = list(title = "Robust Copy Number Metric"),
      showlegend = TRUE
    )
  }
  
  # Package our results nicely for the markdown chunks below
  return(list(
    stats_summary = data.frame(
      Metric = c("Control Region Mean", "Control Region SD", "Significant Bins (Raw)", "Significant Bins (BH Adjusted)"),
      Value  = c(round(baseline_mean, 4), round(baseline_sd, 4), raw_significant_bins, adj_significant_bins)
    ),
    segments_table = significant_segments,
    plot = interactive_map
  ))
}
```


```{r run-cnv-pipeline}
# Boundaries for our flanking negative control zone (known normal region)
control_start_mb <- 25e6
control_end_mb   <- 28e6

# 1. Apply a rolling median window (k=5) to eliminate single-bin noise
smoothed_ratio <- runmed(d_hat, k = 5)

# 2. Transform to log2 space to stabilize variance and center normal cells at 0
# We add a tiny offset (eps) to prevent log2(0) if any bins are completely empty
eps_log <- 1e-4
robust_log2_statistic <- log2(smoothed_ratio + eps_log)

# Execute our generic pipeline using this robust log2 statistic
cnv_results <- find_cnv_regions(
  data_df = df_t, 
  signal_vector = robust_log2_statistic, 
  ctrl_start = control_start_mb, 
  ctrl_end = control_end_mb, 
  alpha = 0.01
)
```


```{r run-cnv-pipeline}
# Boundaries for our flanking negative control zone
control_start_mb <- 25e6
control_end_mb   <- 28e6

# Group nearby cells together using a rolling median window to cut single-cell noise
# Using k = 5 implements the required small-group robust smoothing factor
robust_smooth_ratio <- runmed(d_hat, k = 5)

# Execute the generic setup function using BH adjustment
cnv_results <- find_cnv_regions(
  data_df = df_t, 
  signal_vector = robust_smooth_ratio, 
  ctrl_start = control_start_mb, 
  ctrl_end = control_end_mb, 
  alpha = 0.01
)
```


```{r show-stats-summary}
# Display background noise parameters and before/after significance counts
knitr::kable(
  cnv_results$stats_summary, 
  caption = "Baseline Control Parameters and False Discovery Rate Correction Comparison"
)
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```


```{r show-final-cnvs}
# Table showing continuous blocks of genomic alterations
knitr::kable(
  cnv_results$segments_table, 
  caption = "Contiguous Chromosomal Segments Flagged with Significant Copy Number Alterations"
)

# Render the interactive zoom/pan coordinate plot canvas
cnv_results$plot