---
title: "HW9"
author: "Group 12 - Shir, Omer & Ido"
date: "16/06/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

------------------------------------------------------------------------

### Paths and Data

```{r setup, message=FALSE, warning=FALSE}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }
if (!require('MASS'))       { install.packages('MASS');       library('MASS') }

source("../utils.R")
source("../lab5/lab5_utils.R")

healthy_file <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
tumor_file   <- "../lab7/data/TCGA-13-0723-01A_lib2_all_chr1.forward"
chr1_file    <- "../lab1/data/chr1_str.rda"
```

```{r load-data, cache=TRUE}
healthy_reads <- fread(healthy_file, col.names = c("Chr", "Loc", "Length"))
tumor_reads   <- fread(tumor_file,   col.names = c("Chr", "Loc", "Length"))
load(chr1_file)
```

```{r bin-and-correct, cache=TRUE}
BIN <- 2500

add_pos <- function(df, beg) {
  df$pos <- beg + (seq_len(nrow(df)) - 1) * BIN + BIN / 2
  df
}

# Two arms around centromere (120-130 Mb excluded), both samples
df_h <- rbind(
  add_pos(bin_data(healthy_reads, chr1, 5000001L,   119990000L, BIN), 5000001L),
  add_pos(bin_data(healthy_reads, chr1, 130000001L, 242000000L, BIN), 130000001L)
)
df_t <- rbind(
  add_pos(bin_data(tumor_reads, chr1, 5000001L,   119990000L, BIN), 5000001L),
  add_pos(bin_data(tumor_reads, chr1, 130000001L, 242000000L, BIN), 130000001L)
)

# Per-sample GC spline + plug-in estimator (lab 8 method b)
eps <- 1
gc_correct <- function(df) {
  nz   <- df$reads > 0
  fit  <- fit_gc_spline(df[nz, ])
  fhat <- pmax(predict(fit, data.frame(gc = df$gc)), 0)
  raw  <- (df$reads + eps) / (fhat + eps)
  raw / median(raw[nz])
}
df_h$cn <- gc_correct(df_h)
df_t$cn <- gc_correct(df_t)

# Two-sample estimator (lab 8 method d): GC-corrected tumor / GC-corrected healthy
df_t$cn_2s <- df_t$cn / df_h$cn

nz_h <- df_h$reads > 0
nz_t <- df_t$reads > 0
nz   <- nz_h & nz_t   # bins with reads in both samples

# alias so rest of code uses df / df$cn referring to working data frame
df <- df_t

cat(sprintf("Total bins: %d  |  both non-zero: %d\n", nrow(df), sum(nz)))
```

------------------------------------------------------------------------

## Part A -- Test Statistic

We need a statistic $T$ that summarises whether a genomic region has more or fewer reads than expected under copy number 1, while being **resistant to single noisy bins** -- meaning one extreme bin should not be able to drive the entire region to significance on its own.

We use the **two-sample GC-corrected estimator** (Lab 8 method d): $\hat{a}_k = \hat{a}^\text{tumor}_k / \hat{a}^\text{healthy}_k$, where each sample is corrected by its own GC spline and median-normalised.
Dividing cancels shared GC bias and library-specific technical effects, leaving a cleaner copy-number signal than single-sample correction alone.

We then take $\log_2(\hat{a}_k)$: under $H_0$ (CN = 1) this is centred at 0, gains are positive, losses are negative, and the $\log_2$ scale has a natural interpretation (a value of 1 means a doubling; $-1$ means a halving).

We considered three candidates, all computed as a rolling aggregate over a window of $W = 11$ bins ($\approx 27.5$ kb):

| Statistic | Formula | Notes |
|----|----|----|
| **S1**: plain log$_2$ | $\log_2(\hat{a}_k)$ -- per bin, no aggregation | Baseline |
| **S2**: trimmed mean | $\overline{\log_2(\hat{a})}_{10\%}$ over window | Hard-discards top/bottom 10% of bins |
| **S3**: Huber location | $\hat{\mu}_\text{Huber}(\log_2(\hat{a}),\, k=1.345)$ over window | Continuously downweights outlier bins |

We evaluated all three against the empirical null (Part B).
S2 and S3 produce noticeably narrower, more compressed null distributions when combined with the local-stability negative control filter: the filter already removes noisy bins, so a further robust aggregation over-compresses the tails.
S1 -- plain $\log_2(\hat{a}_k)$ -- keeps the most natural spread.
We carry all three forward to Part C and defer the final choice until we see their detection performance on the full chromosome; the "resistance to single noisy bins" requirement is in any case addressed by the negative control selection (Part B) rather than by the statistic alone.

```{r define-statistics}
W    <- 11
HALF <- (W - 1) / 2

log2_cn <- log2(pmax(df$cn_2s, 1e-6))   # log2 of two-sample a-hat

# Rolling window helper (kept for S2/S3 comparison plots)
roll_stat <- function(x, fun) {
  out <- rep(NA_real_, length(x))
  for (i in seq_along(x)) {
    idx <- max(1, i - HALF) : min(length(x), i + HALF)
    sub <- x[idx][nz[idx]]
    if (length(sub) >= 5) out[i] <- fun(sub)
  }
  out
}

df$S1 <- log2_cn
df$S2 <- roll_stat(log2_cn, function(x) mean(x, trim = 0.10))
df$S3 <- roll_stat(log2_cn, function(x) huber(x, k = 1.345)$mu)
```

```{r stat-distributions, fig.width=11, fig.height=4, dev="png", dpi=180}
par(mfrow = c(1, 3), mar = c(4, 3.5, 2.5, 1))
stats <- list(
  list(v = df$S1[nz], lab = "S1: log2(a-hat)  (per bin)"),
  list(v = df$S2[nz & !is.na(df$S2)], lab = "S2: trimmed mean of log2(a-hat)"),
  list(v = df$S3[nz & !is.na(df$S3)], lab = "S3: Huber of log2(a-hat)")
)
cols <- c("#555555", "#56B4E9", "#009E73")

for (i in seq_along(stats)) {
  v    <- stats[[i]]$v
  v    <- pmin(pmax(v, -2), 2)
  xlim <- c(-2, 2)
  xseq <- seq(xlim[1], xlim[2], length.out = 300)
  h    <- hist(v, breaks = 100, plot = FALSE)
  plot(h, freq = FALSE, col = adjustcolor(cols[i], 0.6), border = NA,
       main = stats[[i]]$lab, xlab = "Statistic value", cex.main = 0.85,
       xlim = xlim)
  abline(v = 0, col = "red", lwd = 2, lty = 2)
  lines(xseq, dnorm(xseq, mean(v, na.rm = TRUE), sd(v, na.rm = TRUE)),
        col = "black", lwd = 2)
  legend("topright", bty = "n", cex = 0.8,
         legend = sprintf("SD=%.3f\nMAD=%.3f", sd(v, na.rm=TRUE), mad(v, na.rm=TRUE)))
}
```

------------------------------------------------------------------------

## Part B -- Negative Control Regions and Null Distribution

**How we select the negative controls.** We need genomic regions where CN = 1 with high confidence, so that the statistic values computed there serve as a direct sample from $H_0$.
We use the paired healthy tissue sample (TCGA-13-0723-10B): a normal blood cell has CN = 1 almost everywhere on chr1, making the healthy genome our natural candidate pool.

We apply the local stability filter (**Strategy B**) directly on the **two-sample ratio** $\log_2(\hat{a}_k)$: a bin is kept as a negative control only if its 11-bin neighbourhood simultaneously has a median near 0 **and** a low local SD.
We set the two cut-offs in interpretable linear-ratio units and then convert to $\log_2$: the local median CN must stay within $\pm 10\%$ of the no-change baseline ($|\tilde{T}| \leq \log_2 1.10 \approx 0.137$), and the local scatter must stay under $\pm 15\%$ (SD $\leq \log_2 1.15 \approx 0.202$).
Filtering on the ratio rather than the healthy sample alone is the correct choice: $H_0$ is "no tumor-specific CN change" (ratio = 1), not "healthy CN = 1 absolutely." This also handles germline CNVs correctly — if both samples carry a duplication at some locus, the ratio is still 1 and the bin is a valid negative control.
Bins at tumor CN-event boundaries are excluded because their neighbours reveal the disruption even if the bin itself lands near 1.
Approximately 37% of non-zero bins pass this filter, giving \~31,000 control bins across both chromosome arms.

**How we model** $H_0$.
The $\log_2(\hat{a}_k)$ values of the selected control bins are a direct sample from the distribution of the statistic when CN = 1 -- we use them directly as the empirical null, without assuming any parametric form.
The histograms and QQ plots below characterise this distribution for each of the three candidate statistics.

**How p-values are derived (Part C).** For any test bin $k$ with statistic value $T_k$, the two-sided empirical p-value is the fraction of null values at least as extreme: $$p_k = \frac{\bigl|\{j \in \text{ctrl} : |T_j| \geq |T_k|\}\bigr|}{|\text{ctrl}|}$$ With \$\sim\$31,000 control bins this resolves p-values down to $\approx 3 \times 10^{-5}$, sufficient for testing at $\alpha = 0.01$ after multiple testing correction.
No normality assumption is required.

```{r neg-ctrl}
# Stability filter runs on the two-sample ratio (log2_cn = log2(tumor/healthy)).
# H0 is "no tumor-specific CN change" (ratio = 1), so the filter should assess
# whether the ratio is locally stable near 0 -- not whether the healthy alone is.
# This also correctly handles germline CNVs: if both samples have CN=2, ratio=1
# and the bin is still a valid negative control.
# Thresholds set in interpretable linear-ratio units, then converted to log2:
#   median within +/-10% of the no-change baseline; local scatter under +/-15%.
med_thresh <- log2(1.10)   # ~0.137
sd_thresh  <- log2(1.15)   # ~0.202

win_med_log <- roll_stat(log2_cn, median)
win_sd_log  <- roll_stat(log2_cn, sd)
stratB <- nz & !is.na(win_med_log) & !is.na(win_sd_log) &
          abs(win_med_log) <= med_thresh & win_sd_log <= sd_thresh

cat(sprintf("Strategy B (local stability): %d bins (%.0f%% of non-zero)\n",
            sum(stratB), 100 * sum(stratB) / sum(nz)))
```

```{r negctrl-scatter, fig.width=9, fig.height=4, dev="png", dpi=180}
# Chromosome-wide scatter: control bins highlighted, excluded in grey
set.seed(42)
idx_all <- which(nz)
s       <- sample(idx_all, min(25000, length(idx_all)))
x_mb    <- df$pos[s] / 1e6
y_cn    <- pmin(log2_cn[s], 3)

col_vec <- ifelse(stratB[s],
                  adjustcolor("#009E73", 0.5),
                  adjustcolor("#999999", 0.2))

par(mar = c(4.5, 4, 2.5, 1), mgp = c(2.2, 0.7, 0))
plot(x_mb, y_cn, pch = 20, cex = 0.25, col = col_vec,
     xlab = "Position (Mb)", ylab = expression(log[2](hat(a)[k])),
     main = "Negative control bins across chr1  (green = control, grey = excluded)",
     ylim = c(-2, 3))
abline(h = 0, col = "#E69F00", lwd = 1.5, lty = 2)
legend("topright", bty = "n", cex = 0.85,
       legend = c(sprintf("Control (%d bins, %.0f%%)",
                          sum(stratB), 100 * sum(stratB) / sum(nz)), "Excluded"),
       col    = c(adjustcolor("#009E73", 0.8), adjustcolor("#999999", 0.6)),
       pch    = 20, pt.cex = 1.2)
```

```{r negctrl-null-hist, fig.width=6, fig.height=4, dev="png", dpi=180}
# Null distribution for S1 on control bins, with normal overlay
null_s1 <- df$S1[stratB & !is.na(df$S1)]
xlim    <- c(-2, 2)
xseq    <- seq(xlim[1], xlim[2], length.out = 300)
h       <- hist(null_s1, breaks = 80, plot = FALSE)
par(mar = c(4, 3.5, 2.5, 1))
plot(h, freq = FALSE, col = adjustcolor("#555555", 0.6), border = NA,
     main = expression("Null distribution: " * log[2](hat(a)[k]) * " on control bins (S1)"),
     xlab = expression(log[2](hat(a)[k]) * " under H0"),
     xlim = xlim, cex.main = 0.9)
abline(v = 0, col = "red", lwd = 1.5, lty = 2)
lines(xseq, dnorm(xseq, mean(null_s1), sd(null_s1)), col = "black", lwd = 2)
legend("topright", bty = "n", cex = 0.85,
       legend = sprintf("n = %d\nMean = %.3f\nSD = %.3f",
                        length(null_s1), mean(null_s1), sd(null_s1)))
```

```{r null-dist-plot, fig.width=13, fig.height=7.3, dev="png", dpi=180}
stat_list_b <- list(S1 = df$S1, S2 = df$S2, S3 = df$S3)
stat_labs_b <- c("S1: log2(a-hat)", "S2: trimmed mean", "S3: Huber")
cols_b      <- c("#555555", "#56B4E9", "#009E73")

par(mfrow = c(2, 3), mar = c(4.5, 4, 3, 1), mgp = c(2.4, 0.8, 0),
    cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)

for (ti in 1:3) {
  mask  <- stratB & !is.na(stat_list_b[[ti]])
  v     <- stat_list_b[[ti]][mask]
  # xlim: 5 SDs each side so shape is always visible, floor at ±0.5
  hw    <- max(5 * sd(v), 0.5)
  xlim  <- c(-hw, hw)
  xseq  <- seq(xlim[1], xlim[2], length.out = 300)
  h     <- hist(v, breaks = 80, plot = FALSE)
  plot(h, freq = FALSE, col = adjustcolor(cols_b[ti], 0.6), border = NA,
       main = sprintf("Null: Strategy B, %s", stat_labs_b[ti]),
       xlab = "Statistic under H0", xlim = xlim)
  abline(v = 0, col = "red", lwd = 1.5, lty = 2)
  lines(xseq, dnorm(xseq, mean(v), sd(v)), col = "black", lwd = 2)
  legend("topright", bty = "n", cex = 1.0,
         legend = sprintf("n=%d\nSD=%.4f\nMAD=%.4f", length(v), sd(v), mad(v)))
}

for (ti in 1:3) {
  mask <- stratB & !is.na(stat_list_b[[ti]])
  v    <- stat_list_b[[ti]][mask]
  qqnorm(v, pch = 20, cex = 0.3, col = adjustcolor(cols_b[ti], 0.4),
         main = sprintf("QQ: Strategy B, %s", stat_labs_b[ti]))
  qqline(v, col = "red", lwd = 2)
}
```

The histograms are shown with a fitted normal overlay (black line) for reference, and the QQ plots assess normality directly.
S2 and S3 are visibly narrower and more compressed than S1, a consequence of applying robust aggregation on top of bins already filtered for stability.
But comparing shapes to a normal curve is only superficial -- what actually matters is whether each statistic produces **calibrated** p-values.
We test this directly below.

### Validating calibration (split-half)

A statistic is only usable if, on data that truly follows $H_0$, its p-values are **Uniform(0,1)** -- equivalently, the realised false-positive rate at level $\alpha$ equals $\alpha$.
We can check this with the control bins alone, no labelled events required:

1.  Split the control bins into **interleaved genomic blocks** (block size 100 bins $\gg$ window $W=11$, so the two sets never share a window). Even blocks build the empirical null; odd blocks are held out as "test" bins.
2.  The held-out bins are themselves CN = 1, so under a correctly calibrated test their two-sided empirical p-values must be uniform.
3.  We measure the deviation from uniform (Kolmogorov--Smirnov $D$) and the realised false-positive rate at $\alpha = 0.01$ and $0.05$.

A statistic whose null is artificially narrow (S2/S3) will make ordinary CN = 1 fluctuations look significant, inflating the false-positive rate above $\alpha$ -- which disqualifies it regardless of how its histogram looks.

```{r calibration}
# Interleaved-block split so null-building and test bins never share a window
ctrl_idx <- which(stratB)
blk       <- 100                          # block size (bins) >> W = 11
block_id  <- ctrl_idx %/% blk
null_ix   <- ctrl_idx[block_id %% 2 == 0] # even blocks -> build null
test_ix   <- ctrl_idx[block_id %% 2 == 1] # odd blocks  -> held-out CN=1 tests

# Add-one two-sided empirical p-value (never zero, valid by construction)
emp_p2 <- function(t_test, t_null) {
  n <- length(t_null)
  at <- abs(t_test)
  vapply(at, function(a) (1 + sum(abs(t_null) >= a)) / (n + 1), numeric(1))
}

p_lists <- list()
calib_rows <- list()
for (ti in seq_along(stat_list_b)) {
  v_null <- stat_list_b[[ti]][null_ix]; v_null <- v_null[is.finite(v_null)]
  v_test <- stat_list_b[[ti]][test_ix]; v_test <- v_test[is.finite(v_test)]
  m <- median(v_null)                     # H0 centre, estimated from the null
  p <- emp_p2(v_test - m, v_null - m)
  p_lists[[ti]] <- p
  calib_rows[[ti]] <- data.frame(
    Statistic = names(stat_list_b)[ti],
    n_test    = length(p),
    FPR_0.01  = round(mean(p < 0.01), 4),
    FPR_0.05  = round(mean(p < 0.05), 4),
    KS_D      = round(suppressWarnings(ks.test(p, "punif")$statistic), 4)
  )
}
calib_tbl <- do.call(rbind, calib_rows)
rownames(calib_tbl) <- NULL
knitr::kable(calib_tbl,
             col.names = c("Statistic", "n (test)", "FPR @ 0.01", "FPR @ 0.05", "KS D"),
             caption = "Calibration on held-out CN=1 bins. Well-calibrated: FPR approx alpha, KS D small.")
```

```{r calibration-plot, fig.width=12, fig.height=5, dev="png", dpi=180}
cols_b <- c("#555555", "#56B4E9", "#009E73")
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1), mgp = c(2.6, 0.8, 0),
    cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.1)

# Full P-P plot: empirical CDF of p-values vs uniform diagonal
plot(NA, xlim = c(0, 1), ylim = c(0, 1),
     xlab = "Nominal p", ylab = "Observed fraction <= p",
     main = "Calibration (P-P plot): full range")
abline(0, 1, col = "grey50", lwd = 2, lty = 2)
for (ti in 1:3) {
  ps <- sort(p_lists[[ti]])
  lines(ps, (seq_along(ps)) / length(ps), col = cols_b[ti], lwd = 2.5)
}
legend("bottomright", bty = "n", cex = 1.1, lwd = 2.5, col = cols_b,
       legend = names(stat_list_b))

# Zoom into the testing-relevant small-p region
plot(NA, xlim = c(0, 0.1), ylim = c(0, 0.1),
     xlab = "Nominal p", ylab = "Observed fraction <= p",
     main = "Calibration: small-p region")
abline(0, 1, col = "grey50", lwd = 2, lty = 2)
abline(v = 0.01, col = "red", lwd = 1.2, lty = 3)
for (ti in 1:3) {
  ps <- sort(p_lists[[ti]])
  lines(ps, (seq_along(ps)) / length(ps), col = cols_b[ti], lwd = 2.5)
}
legend("topleft", bty = "n", cex = 1.1, lwd = 2.5, col = cols_b,
       legend = names(stat_list_b))
```

**Result.** All three statistics are calibrated: every realised false-positive rate sits at or below its nominal $\alpha$, and the P--P curves track the diagonal.
This is expected -- empirical p-values are self-calibrating by construction, because each statistic is ranked against *its own* null, so the over-compression of S2/S3 cancels between the null and the test bins.
The narrow histograms of S2/S3 therefore do **not** cause false positives; superficial non-normality was a red herring.

The check still discriminates, though.
**S1 is the best calibrated** (KS $D \approx 0.009$, with FPR almost exactly $0.01$ and $0.05$), tracking the uniform diagonal nearly perfectly.
S2 and S3 are slightly **conservative** -- their realised FPR falls *below* $\alpha$ and their P--P curves bow under the diagonal -- because the heavy aggregation discretises the statistic and inflates ties.
Conservatism is safe but costs power: at a fixed $\alpha$, a conservative test detects fewer true events.
Calibration alone thus does not eliminate any candidate; it ranks S1 marginally best and sets up the decisive test -- **detection power** -- which we examine next.

------------------------------------------------------------------------

## External sources statement

We used AI (Claude) to design the negative control selection strategy (exploratory analysis comparing MAD-3 vs local stability filters) and to structure this document.
Statistical choices (window width, stability thresholds) and interpretation are our own.
