---
title: "Detecting a Copy-Number Change at chr1:28.5 Mb Despite GC Bias"
subtitle: "Statistics Lab 52568 — Final Presentation"
author: "Group 12 — Shir Tsarfaty, Omer Sutovsky, Ido Ravid"
date: "`r format(Sys.Date(), '%d/%m/%y')`"
output:
  beamer_presentation:
    latex_engine: xelatex
    theme: "metropolis"
    slide_level: 2
    fig_caption: true
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dev = "png", dpi = 200,
                      cache = TRUE)
```

```{r paths_and_data, include=FALSE}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }

source("../utils.R")
source("../lab5/lab5_utils.R")

healthy_file <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
tumor_file   <- "../lab7/data/TCGA-13-0723-01A_lib2_all_chr1.forward"
chr1_file    <- "../lab1/data/chr1_str.rda"

# Converged choices (see presentation_content_map.md)
BIN <- 2500
```

```{r load_data, include=FALSE}
healthy_reads <- fread(healthy_file, col.names = c("Chr", "Loc", "Length"))
tumor_reads   <- fread(tumor_file,   col.names = c("Chr", "Loc", "Length"))
load(chr1_file)
```

<!-- =====================================================================
  PRESENTATION SCAFFOLD — one slide per "##", ~1-2 per stage.
  Content intentionally left as TODO placeholders; fill after deciding:
    (1) single-sample (b) vs two-sample (d) headline method
    (2) which figures to regenerate (see Gaps in presentation_content_map.md)
  Audit landmines (see AUDIT_findings.md):
    - do NOT quote Lab 6 16/84 variance split (double-counted)
    - present t-stat ranking WITH the pmin(y,5) capping caveat
===================================================================== -->

# Background

## The consultant's request

TODO — the request, the data received (`-10B` healthy / `-01A` tumor `.forward`),
external data (reference chr1 for GC).

## The uncorrected region (~28.5 Mb)

TODO — single-panel method-(a) plot, 25–30 Mb. "The event is buried in noise."

```{r bg_uncorrected_plot, eval=FALSE}
# REGENERATE: single-panel uncorrected CN, 25-30 Mb (extract from lab8 qa-plot)
```

# A — Statistical Model

## From fragment counts to copy number

TODO — model equation $Y_k \sim \text{Pois}(\lambda_k),\ \lambda_k = a_k \cdot f(\text{gc}_k) \cdot \epsilon_k$,
each term annotated; VMR $\approx 14.7$ as the motivation that a single $\lambda$ fails.

# B — Modeling the GC Effect

## GC vs. fragment count + spline fit

TODO — `lab7.Rmd:q1-gc-plots` (reuse); cubic B-spline; $r \approx 0.84$;
$R^2$ healthy 0.629 / tumor 0.484.

```{r gc_spline, eval=FALSE}
# REUSE or refit: fit_gc_spline() per sample on non-zero bins
```

## Why fit the spline separately per sample

TODO — overlaid healthy/tumor splines (`lab7.Rmd:q1-compare-curves`).
*(Can merge with previous slide.)*

# C — The Estimator

## Plug-in estimator and the two-sample ratio

TODO — single-sample $\hat a_k = (Y_k/\hat f(\text{gc}_k))\cdot \hat M$;
stabilized $(Y_k+c)/(\hat f+c)$, $c=1$; two-sample ratio (shared noise $\gamma_k$ cancels).

# D — Success Measure

## Quantifying separation of the event

TODO — Welch t-test, event 28–29 Mb vs background 25–28 $\cup$ 29–30 Mb.
**Caveat to address:** `pmin(y,5)` capping affects the ranking.

# E — Corrected Region & Results

## Corrected vs. uncorrected at 28.5 Mb

TODO — single-region before/after overlay (regenerate from `lab8.Rmd:qa-overlay`, 25-30 Mb only).

```{r corrected_overlay, eval=FALSE}
# REGENERATE: single-region (25-30 Mb) overlay, methods (a) and (d) emphasized
```

## The numbers

TODO — Welch t-stat table (re-run `lab8.Rmd:qb-metric` for exact values);
headline: method (d) best, (b) second. Takeaway + caveat.
