---
title: "HW1"
author: "Group Name"
date: "30/03/26"
output:
  html_document:
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

Please use this block to write the paths for all the files you are loading.
If you have a function file, please add it to the submission and call it using a relative path.

```{r paths & global vars}
str1_file = "data/chr1_str.rda"
beg_region = 50000000
end_region = 70000000 - 1

BASES = c("A","C","G","T")

CELL_SIZE=1000
```

#### Run Once - Creating an RDA file from FA file

```{r .rda file creation}
dir_name = "data"
fa_file = sprintf("%s/%s", dir_name, "chr1.fa")

# read as as string
lines = readLines(fa_file)
chr1 = paste(lines[!grepl("^>", lines)], collapse = "")

#save to RDA
save(chr1, file = sprintf("%s/%s", dir_name, "chr1_str.rda"))
```

## Introduction

In the first Lab we are concerned with getting basic statistics about the nucleo-bases in the first chromsome.
We aim to see the distribution of the bases and perhaps gain some insight about recurring patterns.

## Part 1

Reading the Chromosome Data

```{r part1, fig.width=5, fig.height=5}

load(str1_file)  
chr1_reg_str = substr(chr1, beg_region, end_region)
chr_str = strsplit(chr1_reg_str, "")[[1]]
```

some preprocessing:

```{r preprocess}
chr_str = toupper(chr_str) # to upper case
chr_str = chr_str[chr_str != "N"] # removing N

```

## Part 2

Defining the Function:

```{r base count function}
base_count = function(chr_vec, cell_size)
{
  cell_count = floor(length(chr_vec) / cell_size) # calc cell count
  counts = matrix(0, nrow = cell_count, ncol = length(BASES)) # init counts matrix
  colnames(counts) = BASES # set base values as col names
  for (i in 1:cell_count) {
    start_bin = 1 + (i-1)*cell_size
    end_bin = i*cell_size
    for (b in BASES) {
      counts[i, b] = sum(chr_vec[start_bin:end_bin] == b)
    }
  }
  return(counts)
}
```

And using it:

```{r counting for cell_size=1000}
base_counts = base_count(chr_str, CELL_SIZE)
```

## Part 3

```{r, create histogram}
par(mfrow = c(2, 2))
for (b in BASES) {
  hist(base_counts[, b], breaks = 30, main = paste("Base", b), xlab = "Count per cell")
}
```

## Part 4

```{r, proportion plots, fig.width=12, fig.height=8}
par(mfrow = c(2, 2))
n_bins = nrow(base_counts)
bin_size = CELL_SIZE

for (b in BASES) {
  plot((2000:n_bins)*bin_size, base_counts[2000:n_bins, b]/bin_size, 
       pch=20, cex=0.5, xlab="Location", ylab="Proportion", main=paste("Base", b))
  abline(h=0.25)
}
```

```{r, combined plot, fig.width=12, fig.height=6}
CELL_SIZE_SMOOTH = 10000 # increasing the cell size to make the graph clearer
base_counts_smooth = base_count(chr_str, CELL_SIZE_SMOOTH) #. recalculating counts
n_bins_smooth = nrow(base_counts_smooth) # recalc n_bins

# creating a new plot with the 4 bases combined
plot((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "A"]/CELL_SIZE_SMOOTH, 
     type="l", col="lightblue", xlab="Location", ylab="Proportion", 
     main="All Bases", ylim=c(0.1, 0.4))
lines((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "T"]/CELL_SIZE_SMOOTH, col="lightcoral")
lines((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "C"]/CELL_SIZE_SMOOTH, col="darkblue")
lines((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "G"]/CELL_SIZE_SMOOTH, col="darkred")
abline(h=0.25, lty=2)
legend("topright", legend=c("A","T","C","G"), 
       col=c("lightblue","lightcoral","darkblue","darkred"), lty=1)
```

We can clearly see the A-T G-C pairs here based on similar frequencies.
we can also see a region in the beginning (around 5.0e\^6) with higher G-C pairs.

## Part 5

```{r, pair scatter plots, fig.width=12, fig.height=10}
par(mfrow = c(2, 4)) # create grid
bases = c("A","C","G","T")

for (i in 1:3) {
  for (j in (i+1):4) { # increasing innner loop so we dont do the same pair again
    if (i != j) {
      plot(base_counts[,i], base_counts[,j], 
           xlab=bases[i], ylab=bases[j], 
           main=paste(bases[i], "vs", bases[j]), pch=20, cex=0.3)
    }
  }
}

```

1.  We can see negative correlations in the non-paired bases: A-C, A-G, C-T, G-T - all have a (somewhat) negative slope, meaning that one is less prevelant when the other one is more.

2.  We can see that the C-G pair is concentrated in the bottom left corner, with genereally less pairs then the others.
    however, it has outliers in the top right, where both bases have higher frequency.
    looking back to the We can also see that the A-T pair is generally centered higher, indicating a higher shared frequency for it.

```{r, finding cells with high G-C frequency}

high_GC_cells = which(base_counts[,"C"] + base_counts[,"G"] > 550)
high_GC_pos = high_GC_cells * CELL_SIZE + beg_region

plot(1:nrow(base_counts), (base_counts[,"C"] + base_counts[,"G"])/CELL_SIZE, 
     type="l", xlab="Cell Index", ylab="GC Proportion",
     main="High GC Regions", ylim=c(0.3, 0.7))
abline(h=0.5, lty=2, col="gray")
abline(v=c(3000, 5500), lty=1, lwd=3, col="darkred")
legend("bottomright", legend="Region 3000-5500", col="darkred", lty=1, lwd=2)

points(high_GC_cells, (base_counts[high_GC_cells,"C"] + base_counts[high_GC_cells,"G"])/CELL_SIZE, 
       col="blue", pch=20, cex=0.8)


```

We can see the high G-C pairs clustered around cells indexed 3000-5500.
The interesting part is the high density of these pairs, which seems abnormal compared to their distribution elsewhere along the chromosome.

## Part 6

As we mentioned several times, we can see a natural pairing of A-T and C-G.
We can see this pairing in several parts:

1\.
similarity of the base frequency histogram

2\.
frequency graph along the chromosome

3\.
scatter plot outliers

## Short summary

In this first lab we used explored several visualizations for the bases found in the first chromosome.
we looked both at their frequency, their correlation and suggested abnormal regions in the chromosome.

## External sources statement

We used AI (Claude) in several places:

1.  To correctly use `grepl` to save the RDA file.
2.  to fix an off-by-one bug in the `base_count` function
3.  to find the function `which` used to filter the bases frequency in part 5

We also used some R documentation for several functions, namely: [legend](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/legend), [which](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/which)
