---
title: "HW1"
author: "Group Name"
date: "30/03/26"
output:
  html_document:
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

Please use this block to write the paths for all the files you are loading.
If you have a function file, please add it to the submission and call it using a relative path.

```{r paths & global vars}
str1_file = "data/chr1_str.rda"
beg_region = 50000000
end_region = 70000000 - 1

BASES = c("A","C","G","T","a","c","g","t","N")
CONF_BASES = c("A","C","G","T")

CELL_SIZE=1000
```

#### Run Once - Creating an RDA file from FA file

```{r .rda file creation}
dir_name = "data"
fa_file = sprintf("%s/%s", dir_name, "chr1.fa")

# read as as string
lines = readLines(fa_file)
chr1 = paste(lines[!grepl("^>", lines)], collapse = "")

#save to RDA
save(chr1, file = sprintf("%s/%s", dir_name, "chr1_str.rda"))
```

## Introduction

In the first Lab we are concerned with getting basic statistics about the nucleo-bases in the first chromsome.
We aim to see the distribution of the bases and perhaps gain some insight about recurring patterns.

## Part 1

Reading the Chromosome Data

```{r part1, fig.width=5, fig.height=5}

load(str1_file)  
chr1_reg_str = substr(chr1, beg_region, end_region)
chr_str = strsplit(chr1_reg_str, "")[[1]]
```

some preprocessing:\
(Omer's note: there is meaning to upper vs lower case letters, we can't remove it) ((in yuval's rmd he mentioned do to convert to capital, Q%A forum is not open yet in order to ask 0_0 )\
( Ido: I thought since he did this as well, removing lowercase is expected. Keeping it makes the analysis more complex and i don't think that's what he intends. I kept it for now so to keep the graphs from looking ugly. Let's wait for his answer) )

```{r preprocess}
chr_str = toupper(chr_str) # to upper case
chr_str = chr_str[chr_str != "N"] # removing N

```

## Part 2

Defining the Function:

```{r base count function}
base_count = function(chr_vec, cell_size)
{
  cell_count = floor(length(chr_vec) / cell_size) # calc cell count
  counts = matrix(0, nrow = cell_count, ncol = length(BASES)) # init counts matrix
  colnames(counts) = BASES # set base values as col names
  for (i in 1:cell_count) {
    start_bin = 1 + (i-1)*cell_size
    end_bin = i*cell_size
    for (b in BASES) {
      counts[i, b] = sum(chr_vec[start_bin:end_bin] == b)
    }
  }
  return(counts)
}
```

And using it:

```{r counting for cell_size=1000}
base_counts = base_count(chr_str, CELL_SIZE)
```

## Part 3

```{r, create histogram}
par(mfrow = c(2, 2))
for (b in CONF_BASES) {
  hist(base_counts[, b], breaks = 30, main = paste("Base", b), xlab = "Count per cell")
}
```

## Part 4

```{r, proportion plots, fig.width=12, fig.height=8}
par(mfrow = c(2, 2))
n_bins = nrow(base_counts)
bin_size = CELL_SIZE

for (b in CONF_BASES) {
  x_vals <- ((2000:n_bins) * bin_size) / 1e6 # convert to millions
  plot(x_vals, base_counts[2000:n_bins, b] / bin_size,
       pch=20, cex=0.5, xlab="Base Index (millions)", ylab="Proportion", main=paste("Base", b))
  abline(h=0.25)
}
```

```{r, combined plot, fig.width=12, fig.height=6}
CELL_SIZE_SMOOTH = 10000 # increasing the cell size to make the graph clearer
base_counts_smooth = base_count(chr_str, CELL_SIZE_SMOOTH) #. recalculating counts
n_bins_smooth = nrow(base_counts_smooth) # recalc n_bins

# creating a new plot with the 4 bases combined
plot((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "A"]/CELL_SIZE_SMOOTH, 
     type="l", col="lightblue", xlab="Location", ylab="Proportion", 
     main="All Bases", ylim=c(0.1, 0.4))
lines((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "T"]/CELL_SIZE_SMOOTH, col="lightcoral")
lines((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "C"]/CELL_SIZE_SMOOTH, col="darkblue")
lines((200:n_bins_smooth)*CELL_SIZE_SMOOTH, base_counts_smooth[200:n_bins_smooth, "G"]/CELL_SIZE_SMOOTH, col="darkred")
abline(h=0.25, lty=2)
legend("topright", legend=c("A","T","C","G"), 
       col=c("lightblue","lightcoral","darkblue","darkred"), lty=1)
```

(Omer's note: here are my suggestion on how to display it, added a running median line with a window of 201 bins = 100k bases to each side. then layed the trends on top of each other and the comparison is easier i think)

```{r, plotting hists with running median}
par(mfrow = c(2, 2))
options(scipen = 999)

n_bins = nrow(base_counts)
bin_size = CELL_SIZE

k <- 201 # window size

for (b in CONF_BASES) {
  x_vals <- ((2000:n_bins) * bin_size) / 1e6
  y_vals <- base_counts[2000:n_bins, b] / bin_size
  
  plot(x_vals, y_vals,
       pch = 20,
       cex = 0.4,
       col = "black",
       xlab = "Base Index (millions)",
       ylab = "Proportion",
       main = paste("Base", b))
  
  # reference line
  abline(h = 0.25, lty = 2)
  
  # running median
  y_runmed <- runmed(y_vals, k = k)
  
  # thick red trend line
  lines(x_vals, y_runmed, col = "red", lwd = 3)
}
```

```{r, comparing runmed of the bases}
# store results
runmed_list <- list()

for (b in CONF_BASES) {
  x_vals <- ((2000:n_bins) * bin_size) / 1e6
  y_vals <- base_counts[2000:n_bins, b] / bin_size
  runmed_list[[b]] <- runmed(y_vals, k = k)
}

# colors for combined plot
base_colors <- c(A = "red", C = "blue", G = "darkgreen", T = "purple")

par(mar = c(5, 4, 4, 8), xpd = TRUE)
plot(x_vals, runmed_list[[CONF_BASES[1]]],
     type = "n",
     xlab = "Base Index (millions)",
     ylab = "Running Median Proportion",
     main = "Running Median Comparison Across Bases",
     ylim = range(unlist(runmed_list), na.rm = TRUE))

# add lines
for (b in CONF_BASES) {
  lines(x_vals, runmed_list[[b]],
        col = base_colors[b],
        lwd = 2)
}

# legend
legend("topright",
       inset = c(-0.2, 0),
       cex = 0.8,
       legend = paste("Base", CONF_BASES),
       col = base_colors[CONF_BASES],
       lwd = 2,
       bty = "n")
```

We can clearly see the A-T G-C pairs here based on similar frequencies.
It is more easily noticed with the running median trend lines.

## Part 5

```{r, pair scatter plots, fig.width=12, fig.height=10}
par(mfrow = c(2, 4)) # create grid
layout(matrix(c(
  3, 2, 1, 0,
  4, 5, 6, 0
), nrow = 2, byrow = TRUE))
bases = c("A","C","G","T")

for (i in 1:3) {
  for (j in (i+1):4) { # increasing innner loop so we dont do the same pair again
    par(pin = c(3, 3))
    if (i != j) {
      plot(base_counts[,i], base_counts[,j], 
           xlab=bases[i], ylab=bases[j], 
           main=paste(bases[i], "vs", bases[j]), pch=20, cex=0.3, asp = 1)
    }
  }
}

```

1.  We can see negative correlations in the non-paired bases: A-C, A-G, C-T, G-T - all have a (somewhat) negative slope, meaning that one is less prevelant when the other one is more. (Omer's note: I dont see the negative slope you are refreing to.) 1.\* We can see all of the pairs have somewhat of a positive correlation. But it is clear that the strongest pairs are A-T and C-G.
2.  We can see that the C-G pair is concentrated in the bottom left corner, with genereally less pairs then the others. however, it has outliers in the top right, where both bases have higher frequency. looking back to the We can also see that the A-T pair is generally centered higher, indicating a higher shared frequency for it.

```{r, finding cells with high G-C frequency}

high_GC_cells = which(base_counts[,"C"] + base_counts[,"G"] > 550)
high_GC_pos = high_GC_cells * CELL_SIZE + beg_region

plot(1:nrow(base_counts), (base_counts[,"C"] + base_counts[,"G"])/CELL_SIZE, 
     type="l", xlab="Cell Index", ylab="GC Proportion",
     main="High GC Regions", ylim=c(0.3, 0.7))
abline(h=0.5, lty=2, col="gray")

# Shade the high GC region
rect(3000, 0.3, 5500, 0.7, col=rgb(1, 0, 0, 0.1), border=NA)

abline(v=c(3000, 5500), lty=1, lwd=2, col="darkred")
legend("bottomright", legend="Region 3000-5500 (53 M-55.5 M)", col="darkred", lty=1, lwd=2, cex = 0.8)


```

(Omer's note: sohuldnt the regoin be between 50-70 mil? i am a bit confused on this part lol)

(Ido: this region i marked is the one that shows the abnormally high GC pairs. I clarified it)

We can see the high G-C pairs clustered around cells indexed 3000-5500, which correspond to the area between 53 M-55.5 M. The interesting part is the high density of these pairs, which seems abnormal compared to their distribution elsewhere along the chromosome.

## Part 6

As we mentioned several times, we can see a natural pairing of A-T and C-G.
We can see this pairing in several parts:

1\.
similarity of the base frequency histogram and running medians

2\.
frequency graph along the chromosome

3\.
scatter plot outliers

## Short summary

In this first lab we used explored several visualizations for the bases found in the first chromosome.
we looked both at their frequency, their correlation and suggested abnormal regions in the chromosome.

## External sources statement

We used AI (Claude) in several places:

1.  To correctly use `grepl` to save the RDA file.
2.  To fix an off-by-one bug in the `base_count` function
3.  To find the function `which` used to filter the bases frequency in part 5

We also used some R documentation for several functions, namely: [legend](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/legend), [which](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/which)
