---
title: "HW4"
author: "Group 12 - Shir, Omer & Ido"
date: "10/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths, packages & global vars}
if (!require('data.table'))  { install.packages('data.table');  library('data.table') }
if (!require('splines'))     { install.packages('splines');     library('splines') }

reads_file  <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
chr1_file   <- "../lab1/data/chr1_str.rda"

beg_region  <- 50000000
end_region  <- 100000000  # 50Mb region as in lab3 Q3
BIN_SIZE    <- 5000
```

### Reading the Data:

```{r loading data}
source("../utils.R")

# Load reads data
chr1_reads <- fread(reads_file, col.names = c("Chr", "Loc", "Length"))

# Load chromosome sequence
load(chr1_file)

# Calculate coverage for the 50Mb region
N <- end_region - beg_region + 1
cov <- count_reads_table(chr1_reads, beg_region, end_region)

# Bin coverage
bin_counts <- sapply(seq(1, N - BIN_SIZE + 1, by = BIN_SIZE), function(i) {
  sum(cov[i:(i + BIN_SIZE - 1)])
})

# Calculate GC content for the same region
seq_region <- substr(chr1, beg_region, end_region)
split_seq  <- strsplit(seq_region, "")[[1]]

gc_binned <- sapply(seq(1, length(split_seq) - BIN_SIZE + 1, by = BIN_SIZE), function(i) {
  sum(split_seq[i:(i + BIN_SIZE - 1)] %in% c("C", "G", "c", "g"))
})

# Align lengths and convert GC to proportions
min_len <- min(length(bin_counts), length(gc_binned))
reads_data <- bin_counts[1:min_len]
gc_data <- gc_binned[1:min_len] / BIN_SIZE

cat(sprintf("Data loaded: %d bins, GC range: %.2f-%.2f\n",
            length(reads_data), min(gc_data), max(gc_data)))
```

## Introduction

In this lab we are fitting a non-parametric (homoscedastic) spline regression, to estimate the reads-GC content relationship.
Our assumption is that the the relationship is continuous and therefore a spline regression is a valid tool for modelling it.
Once we have this function, we can "remove" it from the actual reads data and analyze the connection between reads and mutation in the chromosome.

## Part 1

### Candidate Models

We fit four different non-parametric spline models:

```{r model-fitting}
# Create prediction grid for all models
gc_grid <- seq(min(gc_data), max(gc_data), length.out = 500)
pred_grid <- data.frame(gc_data = gc_grid)

# Store models and their info
models <- list()
preds_list <- list()
model_info <- data.frame()

# ===== MODEL 1:  cubic, 3 knots at quartiles =====
knots_1 <- quantile(gc_data, probs = c(0.25, 0.5, 0.75))
model1 <- lm(reads_data ~ bs(gc_data, knots = knots_1, degree = 3))
preds_1 <- predict(model1, pred_grid)

models[[1]] <- model1
preds_list[[1]] <- preds_1
model_info <- rbind(model_info, data.frame(
  name = "model1",
  description = "3 quartile knots, cubic"
))

# ===== MODEL 2: cubic, 3 knots at 1/3 and 2/3 precentiles  =====
knots_2 <- quantile(gc_data, probs = c(1/3, 2/3))
model2 <- lm(reads_data ~ bs(gc_data, knots = knots_2, degree = 3))
preds_2 <- predict(model2, pred_grid)

models[[2]] <- model2
preds_list[[2]] <- preds_2
model_info <- rbind(model_info, data.frame(
  name = "model2",
  description = "2 knots at 1/3 and 2/3, cubic"
))

# ===== MODEL 3: cubic, 1 knot at median =====
knots_3 <- median(gc_data)
model3 <- lm(reads_data ~ bs(gc_data, knots = knots_3, degree = 3))
preds_3 <- predict(model3, pred_grid)

models[[3]] <- model3
preds_list[[3]] <- preds_3
model_info <- rbind(model_info, data.frame(
  name = "model3",
  description = "1 median knot, cubic"
))

# ===== MODEL 4: quadratic,  3 knots at quartiles =====
model4 <- lm(reads_data ~ bs(gc_data, knots = knots_1, degree = 2))
preds_4 <- predict(model4, pred_grid)

models[[4]] <- model4
preds_list[[4]] <- preds_4
model_info <- rbind(model_info, data.frame(
  name = "model4",
  description = "3 quartile knots, quadratic (degree 2)"
))

cat("=== Models Fitted ===\n")
print(model_info)
```

### Question 1: Individual model plots

```{r q1-model-comparison, fig.width=12, fig.height=14}
par(mfrow = c(4, 1), mar = c(4, 4, 2, 1))

model_colors <- c("red", "blue", "green", "purple")
model_names <- c("model1", "model2", "model3", "model4")

for (i in 1:4) {
  plot(gc_data, reads_data,
       col = rgb(0, 0, 0, 0.2),
       pch = 20, cex = 0.4,
       xlab = "GC Content (proportion)",
       ylab = "Reads per bin",
       main = paste0(model_names[i], ": ", model_info$description[i]),
       ylim = c(0, 400))

  # Add model line
  lines(gc_grid, preds_list[[i]], col = model_colors[i], lwd = 3)

  # Add reference lines for knots (if applicable)
  if (i == 1) {
    abline(v = knots_1, col = model_colors[i], lty = 2, lwd = 2.5)
  }
  if (i == 2) {
    abline(v = knots_2, col = model_colors[i], lty = 2, lwd = 2.5)
  }
  if (i == 3) {
    abline(v = knots_3, col = model_colors[i], lty = 2, lwd = 2.5)
  }
  if (i == 4) {
    abline(v = knots_1, col = model_colors[i], lty = 2, lwd = 2.5)
  }
}

par(mfrow = c(1, 1))
```

We see generally that all 4 models fit well the general pattern of the data, with the extrapolation part around 0.6 GC and above being different between the different models.

### Question 2: Residual analysis (all models)

```{r q2-residuals-comparison, fig.width=12, fig.height=14}
par(mfrow = c(4, 1), mar = c(4, 4, 2, 1))

residuals_list <- lapply(models, residuals)

for (i in 1:4) {
  residuals_i <- residuals_list[[i]]

  plot(gc_data, residuals_i,
       col = rgb(0, 0, 0, 0.5),
       pch = 20, cex = 0.4,
       xlab = "GC Content (proportion)",
       ylab = "Residuals",
       main = paste0(model_names[i], ": Residuals vs GC"),
       ylim = c(-200, 200))
  abline(h = 0, col = "red", lty = 2, lwd = 1.5)
}

par(mfrow = c(1, 1))
```

The residual graphs between the different models are largely identical, and it's hard for us to spot visually any meaningful difference.

### Question 3: Model quality metrics comparison

```{r q3-metrics-table}
# Calculate metrics for all models
metrics_table <- data.frame()

for (i in 1:4) {
  mod <- models[[i]]
  resid <- residuals_list[[i]]

  rmse <- sqrt(mean(resid^2))
  r2 <- summary(mod)$r.squared
  adj_r2 <- summary(mod)$adj.r.squared
  aic <- AIC(mod)
  bic <- BIC(mod)

  # Heteroscedasticity check
  gc_low <- gc_data < quantile(gc_data, 0.25)
  gc_high <- gc_data > quantile(gc_data, 0.75)
  sd_ratio <- sd(resid[gc_high]) / sd(resid[gc_low])

  metrics_table <- rbind(metrics_table, data.frame(
    Model = model_names[i],
    RMSE = round(rmse, 3),
    R2 = round(r2, 4),
    Adj_R2 = round(adj_r2, 4),
    SD_ratio = round(sd_ratio, 2),
    stringsAsFactors = FALSE
  ))
}

cat("\n=== Model Comparison Table ===\n")
print(metrics_table)

```

We tried to tweak the amount of knots and the degree of the splines, it mostly had little effect.
The parameter that impacted most was the amount of knots, which makes sense that will allow for better fitting of the data.

We calculated the standard deviation of residuals, in the top and bottom quartiles of GC content and calculated their ratio.
We expected the value to be close to 1, since assuming homoscadacity the variance should be more or less uniform across different GC values.
The 2.25 SD ratio indicates we have heteroscedastic data distribution, and we might need to attend to that in later labs.

## Part 2

### Question 4 : Prediction Function

```{r q4-Prediction Function}
# Prepare the data frame first
full_df <- data.frame(reads = reads_data, gc = gc_data) 

# Define the prediction function
evaluate_spline <- function(train_data, val_data) {
  # Fit spline model on training data (using Model 1 parameters)
  train_knots <- quantile(train_data$gc, probs = c(0.25, 0.5, 0.75))
  fit <- lm(reads ~ bs(gc, knots = train_knots, degree = 3), data = train_data)

  # Predict on validation data
  preds <- predict(fit, newdata = val_data)

  # Calculate RMSE
  rmse_val <- sqrt(mean((val_data$reads - preds)^2))
  return(rmse_val)
}

# Create a split for visual check
set.seed(123)
idx_viz <- sample(seq_len(nrow(full_df)), size = floor(0.5 * nrow(full_df)))
train_viz <- full_df[idx_viz, ]
val_viz   <- full_df[-idx_viz, ]

# Train model on the training set and plot against validation set
viz_knots <- quantile(train_viz$gc, probs = c(0.25, 0.5, 0.75))
fit_viz <- lm(reads ~ bs(gc, knots = viz_knots, degree = 3), data = train_viz)

# Create a smooth line for the prediction
gc_grid_viz <- seq(min(full_df$gc), max(full_df$gc), length.out = 300)
line_preds <- predict(fit_viz, newdata = data.frame(gc = gc_grid_viz))

# The Plot: GC vs. Coverage 
plot(val_viz$gc, val_viz$reads, 
     col = rgb(0, 0, 0, 0.2), pch = 20, cex = 0.5,
     xlab = "GC Content (proportion)", ylab = "Reads (coverage)",
     main = "Visual Validation: Model trained on Train Set vs. Validation Points")
lines(gc_grid_viz, line_preds, col = "red", lwd = 3)
legend("topleft", legend=c("Validation Data", "Spline Fit (Train)"), 
       col=c("gray", "red"), pch=c(20, NA), lwd=c(NA, 3))
```

we can see the model has a good grasp on the validation data points, at least visually.
We do see that in the upper precentiles there is some distance between the model prediction and actual data, but for the most part it looks correct.

### Question 5: RMSE vs. Training Sample Size

```{r q5-learning curve}

# Define a range of sample sizes to test (from 10 up to 5000)
sample_sizes <- seq(10, 5000, length.out = 20)
results <- data.frame(n = sample_sizes, rmse = NA)

# Loop through sample sizes to see how performance changes
for(i in seq_along(sample_sizes)) {
  n <- floor(sample_sizes[i])
  # Take a random sample of size n from our training pool
  temp_train <- train_viz[sample(nrow(train_viz), n), ]
  
  # Evaluate using the function from section (d)
  results$rmse[i] <- evaluate_spline(temp_train, val_viz)
}

# Plot the relationship between sample size and prediction quality
plot(results$n, results$rmse, type = "b", pch = 19, col = "darkblue",
     xlab = "Training Sample Size", 
     ylab = "Validation RMSE",
     main = "Learning Curve: RMSE vs. Training Size",
     ylim = c(0, 100))
grid()
```

We saw that the RMSE drops sharply from about 700 at the start and levels off to 26 almost immediately.
This shows the model learns the main trend with very little data, and adding more observations doesn't significantly improve its accuracy.

## Short summary

We fitted 4 spline regression models with different parameters, and looked at how well each model fitted.
We generally didn't see much diffnece between the models we selected.
We did manage to notice that the model ability to explain the data is capped at \~0.75 R2.
A possible explanation we came up with was the Homoscadacity assumption is incorrect, as we shown with the SD ratio statistic.

In the second part, our validation tests showed that the RMSE starts very high when the training sample is tiny, but it drops sharply almost immediately.
We saw that the model picks up the main GC-coverage trend very quickly.
Once it has around 500 observations, the accuracy plateaus, showing that it’s efficient at capturing the biological signal even from small datasets.

## External sources statement

We used AI (Claude) to:

1.  calculate SD ratio to verify homoscedacity

2.  stack the graphs correctly in q1, q2

3.  fix data loading error in bin size calculation.

4.  Adjust the learning curve scale for better visualization.
