---
title: "HW4"
author: "Group 12 - Shir, Omer & Ido"
date: "10/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths, packages & global vars}
if (!require('data.table'))  { install.packages('data.table');  library('data.table') }
if (!require('splines'))     { install.packages('splines');     library('splines') }

reads_file  <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
chr1_file   <- "../lab1/data/chr1_str.rda"

beg_region  <- 50000000
end_region  <- 100000000  # 50Mb region as in lab3 Q3
BIN_SIZE    <- 5000
```

### Reading the Data:

```{r loading data}
source("../utils.R")

# Load reads data
chr1_reads <- fread(reads_file, col.names = c("Chr", "Loc", "Length"))

# Load chromosome sequence
load(chr1_file)

# Calculate coverage for the 50Mb region
N <- end_region - beg_region + 1
cov <- count_reads_table(chr1_reads, beg_region, end_region)

# Bin coverage
bin_counts <- sapply(seq(1, N - BIN_SIZE + 1, by = BIN_SIZE), function(i) {
  sum(cov[i:(i + BIN_SIZE - 1)])
})

# Calculate GC content for the same region
seq_region <- substr(chr1, beg_region, end_region)
split_seq  <- strsplit(seq_region, "")[[1]]

gc_binned <- sapply(seq(1, length(split_seq) - BIN_SIZE + 1, by = BIN_SIZE), function(i) {
  sum(split_seq[i:(i + BIN_SIZE - 1)] %in% c("C", "G", "c", "g"))
})

# Align lengths and convert GC to proportions
min_len <- min(length(bin_counts), length(gc_binned))
reads_data <- bin_counts[1:min_len]
gc_data <- gc_binned[1:min_len] / BIN_SIZE

cat(sprintf("Data loaded: %d bins, GC range: %.2f-%.2f\n",
            length(reads_data), min(gc_data), max(gc_data)))
```

## Introduction
In this lab we are fitting a non-parametric (homoscedastic) spline regression,
to estimate the reads-GC content relationship. Our assumption is that the the relationship is continuous and therefore a spline regression is a valid tool for modelling it. Once we have this function, we can "remove" it from the actual reads data and analyze the connection between reads and mutation in the chromosome.


## Part 1

### Candidate Models

We fit four different non-parametric spline models:

```{r model-fitting}
# Create prediction grid for all models
gc_grid <- seq(min(gc_data), max(gc_data), length.out = 500)
pred_grid <- data.frame(gc_data = gc_grid)

# Store models and their info
models <- list()
preds_list <- list()
model_info <- data.frame()

# ===== MODEL 1:  cubic, 3 knots at quartiles =====
knots_1 <- quantile(gc_data, probs = c(0.25, 0.5, 0.75))
model1 <- lm(reads_data ~ bs(gc_data, knots = knots_1, degree = 3))
preds_1 <- predict(model1, pred_grid)

models[[1]] <- model1
preds_list[[1]] <- preds_1
model_info <- rbind(model_info, data.frame(
  name = "model1",
  description = "3 quartile knots, cubic"
))

# ===== MODEL 2: cubic, 3 knots at 1/3 and 2/3 precentiles  =====
knots_2 <- quantile(gc_data, probs = c(1/3, 2/3))
model2 <- lm(reads_data ~ bs(gc_data, knots = knots_2, degree = 3))
preds_2 <- predict(model2, pred_grid)

models[[2]] <- model2
preds_list[[2]] <- preds_2
model_info <- rbind(model_info, data.frame(
  name = "model2",
  description = "2 knots at 1/3 and 2/3, cubic"
))

# ===== MODEL 3: cubic, 1 knot at median =====
knots_3 <- median(gc_data)
model3 <- lm(reads_data ~ bs(gc_data, knots = knots_3, degree = 3))
preds_3 <- predict(model3, pred_grid)

models[[3]] <- model3
preds_list[[3]] <- preds_3
model_info <- rbind(model_info, data.frame(
  name = "model3",
  description = "1 median knot, cubic"
))

# ===== MODEL 4: quadratic,  3 knots at quartiles =====
model4 <- lm(reads_data ~ bs(gc_data, knots = knots_1, degree = 2))
preds_4 <- predict(model4, pred_grid)

models[[4]] <- model4
preds_list[[4]] <- preds_4
model_info <- rbind(model_info, data.frame(
  name = "model4",
  description = "3 quartile knots, quadratic (degree 2)"
))

cat("=== Models Fitted ===\n")
print(model_info)
```

### Question 1: Individual model plots

```{r q1-model-comparison, fig.width=12, fig.height=14}
par(mfrow = c(4, 1), mar = c(4, 4, 2, 1))

model_colors <- c("red", "blue", "green", "purple")
model_names <- c("model1", "model2", "model3", "model4")

for (i in 1:4) {
  plot(gc_data, reads_data,
       col = rgb(0, 0, 0, 0.2),
       pch = 20, cex = 0.4,
       xlab = "GC Content (proportion)",
       ylab = "Reads per bin",
       main = paste0(model_names[i], ": ", model_info$description[i]),
       ylim = c(0, 400))

  # Add model line
  lines(gc_grid, preds_list[[i]], col = model_colors[i], lwd = 3)

  # Add reference lines for knots (if applicable)
  if (i == 1) {
    abline(v = knots_1, col = model_colors[i], lty = 2, lwd = 2.5)
  }
  if (i == 2) {
    abline(v = knots_2, col = model_colors[i], lty = 2, lwd = 2.5)
  }
  if (i == 3) {
    abline(v = knots_3, col = model_colors[i], lty = 2, lwd = 2.5)
  }
}

par(mfrow = c(1, 1))
```

We se generally that all 4 models fit well the general pattern of the data, with the extraploation part around 0.6 GC and above being differnt between the different models.

### Question 2: Residual analysis (all models)

```{r q2-residuals-comparison, fig.width=12, fig.height=14}
par(mfrow = c(4, 1), mar = c(4, 4, 2, 1))

residuals_list <- lapply(models, residuals)

for (i in 1:4) {
  residuals_i <- residuals_list[[i]]

  plot(gc_data, residuals_i,
       col = rgb(0, 0, 0, 0.5),
       pch = 20, cex = 0.4,
       xlab = "GC Content (proportion)",
       ylab = "Residuals",
       main = paste0(model_names[i], ": Residuals vs GC"),
       ylim = c(-200, 200))
  abline(h = 0, col = "red", lty = 2, lwd = 1.5)
}

par(mfrow = c(1, 1))
```
The residual graphs between the different models are largely identical, and it's hard for us to spot visually any meaningful difference. 

### Question 3: Model quality metrics comparison

```{r q3-metrics-table}
# Calculate metrics for all models
metrics_table <- data.frame()

for (i in 1:4) {
  mod <- models[[i]]
  resid <- residuals_list[[i]]

  rmse <- sqrt(mean(resid^2))
  r2 <- summary(mod)$r.squared
  adj_r2 <- summary(mod)$adj.r.squared
  aic <- AIC(mod)
  bic <- BIC(mod)

  # Heteroscedasticity check
  gc_low <- gc_data < quantile(gc_data, 0.25)
  gc_high <- gc_data > quantile(gc_data, 0.75)
  sd_ratio <- sd(resid[gc_high]) / sd(resid[gc_low])

  metrics_table <- rbind(metrics_table, data.frame(
    Model = model_names[i],
    RMSE = round(rmse, 3),
    R2 = round(r2, 4),
    Adj_R2 = round(adj_r2, 4),
    SD_ratio = round(sd_ratio, 2),
    stringsAsFactors = FALSE
  ))
}

cat("\n=== Model Comparison Table ===\n")
print(metrics_table)

```
We tried to tweak the amount of knots and the degree of the splines, it mostly had little effect. 
The parameter that impacted most was the amount of knots, which makes sense that will allow for better fitting of the data. 

We calculated the standard deviation of residuals, in the top and bottom quartiles of GC content and calculated their ratio. We expected the value to be close to 1, since assuming homoscadacity the varaince should be more or less uniform across different GC values. The 2.25 SD ratio indicates we have heteroscedastic data distribution, and we might need to attend to that in later labs.

## Part 2
 @add
 
 
## Short summary
We fitted 4 spline regression models with different parameters, and looked at how well each model fitted. We generally didn't see much diffnece between the models we selected. We did manage to notice that the model ability to explain the data is capped at ~0.75 R2. A possible explanation we came up with was the Homoscadacity assumption is incorrect, as we shown with the SD ratio statistic.


 @add
 
 
## External sources statement

We used AI (Claude) to:
1. calculate SD ratio to verify homoscedacity
2. stack the graphs correctly in q1, q2
3. fix data loading error in bin size calculation. 
 @add
