---
title: "HW5"
author: "Group 12 - Shir, Omer & Ido"
date: "19/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths-and-packages, message=FALSE, warning=FALSE}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }
if (!require('shiny'))      { install.packages('shiny');      library('shiny') }

reads_file <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
chr1_file  <- "../lab1/data/chr1_str.rda"

BIN_SIZE   <- 5000
beg_region <- 50000000
end_region <- 100000000
```

### Reading the Data

```{r load-data, cache=TRUE}
source("../utils.R")
source("lab5_utils.R")

chr1_reads <- fread(reads_file, col.names = c("Chr", "Loc", "Length"))
load(chr1_file)

# Main analysis region (50–100 Mb, 5000 bp bins)
df_main <- bin_data(chr1_reads, chr1, beg_region, end_region, BIN_SIZE)

cat(sprintf("Loaded %d bins | GC range: %.3f – %.3f | reads range: %d – %d\n",
            nrow(df_main), min(df_main$gc), max(df_main$gc),
            min(df_main$reads), max(df_main$reads)))
```

## Introduction

Building on labs 3 and 4, this lab investigates the GC-coverage relationship in greater depth.
We compare the predictive power of G, C, and G+C separately, examine whether the relationship holds consistently across chromosome 1, explore the effect of bin size on correlation strength, and fit Poisson regression models — motivated by the overdispersion we measured previously (VMR ≈ 14.7 in lab3, SD_ratio ≈ 2.25 in lab4).

------------------------------------------------------------------------

## Section a — Does C+G explain coverage better than G or C alone?

```{r section-a-models}
# Four nested models
mod_g   <- lm(reads ~ g_prop,                   data = df_main)
mod_c   <- lm(reads ~ c_prop,                   data = df_main)
#mod_gc  <- lm(reads ~ g_prop + c_prop,           data = df_main)
#mod_gci <- lm(reads ~ g_prop * c_prop,           data = df_main)  # includes interaction g_prop:c_prop
mod_gc_sum <- lm(reads ~ gc, data = df_main)

models_a  <- list(mod_g, mod_c, mod_gc_sum)
labels_a  <- c("~G", "~C", "~G+C")

# Summary table
coef_table <- do.call(rbind, lapply(seq_along(models_a), function(i) {
  mod  <- models_a[[i]]
  cf   <- coef(mod)
  ci   <- confint(mod)
  r2   <- summary(mod)$r.squared
  # one row per model: just show R² and the non-intercept coefficients as a string
  data.frame(
    Model    = labels_a[i],
    R2       = round(r2, 4),
    Coefs    = paste(sprintf("%s=%.2f [%.2f,%.2f]",
                             names(cf)[-1], cf[-1], ci[-1,1], ci[-1,2]),
                     collapse = "  |  "),
    stringsAsFactors = FALSE
  )
}))

cat("\n=== Model comparison: R² and coefficients ===\n")
print(coef_table, row.names = FALSE)
```

```{r code}
models_main <- list(
  mod_g,
  mod_c,
  mod_gc_sum
)

labels_main <- c(
  "G alone",
  "C alone",
  "GC = G + C"
)

predictors_main <- list(
  df_main$g_prop,
  df_main$c_prop,
  df_main$gc
)

y <- df_main$reads

baseline_medae <- median(abs(y - median(y)))

model_compare <- do.call(rbind, lapply(seq_along(models_main), function(i) {
  
  mod <- models_main[[i]]
  y_hat <- predict(mod)
  resid <- y - y_hat
  
  r2 <- summary(mod)$r.squared
  
  spearman <- cor(predictors_main[[i]], y, method = "spearman")
  
  medae <- median(abs(resid))
  
  #robust_r2 <- 1 - medae / baseline_medae
  
  data.frame(
    Model = labels_main[i],
    R2 = round(r2, 4),
    Spearman = round(spearman, 4),
    Median_Abs_Error = round(medae, 2)
    #Robust_R2 = round(robust_r2, 4)
  )
}))

print(model_compare)
```

```{r section-a-gcskew, fig.width=7, fig.height=6}
set.seed(42)
keep         <- sample(nrow(df_main), size = min(2000, nrow(df_main)))
df_sub       <- df_main[keep, ]
ord          <- order(df_sub$reads)
reads_sorted <- df_sub$reads[ord]
reads_scaled <- (reads_sorted - min(reads_sorted)) / diff(range(reads_sorted))
pal          <- colorRampPalette(c("#440154", "#21908C", "#FDE725"))(100)
point_cols   <- pal[pmax(1, ceiling(reads_scaled * 100))]

plot(df_sub$g_prop[ord], df_sub$c_prop[ord],
     col  = point_cols, pch = 20, cex = 0.8,
     xlab = "G content per bin",
     ylab = "C content per bin",
     main = "GC Skew: G% vs C% (color = read depth)",
     xlim = c(0, 0.4), ylim = c(0, 0.4))

abline(a = 0, b = 1, col = "black", lwd = 2, lty = 2)

mid_x <- mean(par("usr")[1:2])

# Gradient colorbar — bottom right
usr <- par("usr")
bar_x0 <- usr[2] - 0.055; bar_x1 <- usr[2] - 0.005
bar_y0 <- usr[3] + 0.005; bar_y1 <- usr[3] + 0.07
n_seg <- 50
ys <- seq(bar_y0, bar_y1, length.out = n_seg + 1)
bar_pal <- colorRampPalette(c("#440154", "#21908C", "#FDE725"))(n_seg)
for (k in seq_len(n_seg))
  rect(bar_x0, ys[k], bar_x1, ys[k+1], col = bar_pal[k], border = NA)
rect(bar_x0, bar_y0, bar_x1, bar_y1, border = "black", lwd = 0.7)
text(bar_x0 - 0.002, bar_y0, "Low\nreads",  adj = c(1, 0),   cex = 0.72)
text(bar_x0 - 0.002, bar_y1, "High\nreads", adj = c(1, 1),   cex = 0.72)
```

The second graph is our attempt to further explain the difference in explainability.
The cloud follows the Chargaff diagonal (G% ≈ C% globally), but notably fans out at higher GC values.
Low-GC bins (purple, low reads) sit tightly on the diagonal — G and C are nearly interchangeable there.
High-GC bins (yellow, high reads) spread away from it: local G% and C% diverge due to transcription-coupled strand asymmetry in gene-rich regions.
This means C carries **independent coverage signal precisely where coverage is highest** — which is why `~G+C` outperforms `~G` alone despite Chargaff.

------------------------------------------------------------------------

## Section b — Is the relationship spatially consistent? (Shiny)

The plot below shows spline model R² (for model1 from lab4) as a function of genomic position across the full chromosome.
Run `shiny::runApp("shiny_app")` for the interactive version with a live slider.

```{r section-b-static, fig.width=12, fig.height=5, cache=TRUE}
chr1_len    <- nchar(chr1)
window_size <- 10000000  # 10 Mb
step_size   <- 5000000   # 5 Mb steps
pos_steps   <- seq(1, chr1_len - window_size, by = step_size)

r2_profile <- sapply(pos_steps, function(beg_w) {
  df_w <- bin_data(chr1_reads, chr1, beg_w, beg_w + window_size - 1, BIN_SIZE)
  if (nrow(df_w) < 20 || var(df_w$gc) < 1e-10) return(NA)
  knots_w <- quantile(df_w$gc, probs = c(0.25, 0.5, 0.75))
  fit_w   <- lm(reads ~ bs(gc, knots = knots_w, degree = 3), data = df_w)
  summary(fit_w)$r.squared
})

pos_mb <- pos_steps / 1e6

plot(pos_mb + 5, r2_profile,
     type = "p", pch = 19, cex = 0.9, col = "steelblue",
     xlim = c(0, chr1_len / 1e6), ylim = c(0, 1),
     xlab = "Chromosome 1 position (Mb)", ylab = "R²  (spline fit, 10 Mb window)",
     main = "GC–coverage model fit across chromosome 1",
     xaxs = "i")

# Outer band: windows that partially overlap the centromere (~10 Mb buffer each side)
rect(110, 0, 140, 1, col = adjustcolor("firebrick", alpha.f = 0.08), border = NA)
# Core centromere
rect(120, 0, 130, 1, col = adjustcolor("firebrick", alpha.f = 0.20), border = NA)
text(125, 0.95, "centromere", col = "firebrick", cex = 1.2, adj = 0.5, font = 2)

abline(h = seq(0.2, 0.8, by = 0.2), col = "gray80", lty = 2, lwd = 0.7)
```

------------------------------------------------------------------------

## Section c — Which bin size gives the strongest relationship?

```{r section-c-static, fig.width=13, fig.height=6, cache=TRUE}
bin_sizes_c <- c(500, 1000, 2000, 5000, 10000, 20000, 50000, 100000, 250000, 500000)

windows_c <- list(
  list(beg = 5000000,   end = 25000000,  label = "Edge  (5–25 Mb)",         col = "#377EB8"),
  list(beg = 60000000,  end = 80000000,  label = "Gene-rich  (60–80 Mb)",   col = "#4DAF4A"),
  list(beg = 115000000, end = 135000000, label = "Centromere  (115–135 Mb)", col = "#E41A1C")
)

r2_all <- matrix(NA, nrow = length(bin_sizes_c), ncol = length(windows_c))
sp_all <- matrix(NA, nrow = length(bin_sizes_c), ncol = length(windows_c))

for (j in seq_along(windows_c)) {
  for (k in seq_along(bin_sizes_c)) {
    df_w <- bin_data(chr1_reads, chr1, windows_c[[j]]$beg, windows_c[[j]]$end, bin_sizes_c[k])
    if (nrow(df_w) < 5 || var(df_w$gc) < 1e-10) next
    mod_w        <- lm(reads ~ gc, data = df_w)
    r2_all[k, j] <- summary(mod_w)$r.squared
    sp_all[k, j] <- spearman_r(df_w$gc, df_w$reads)
  }
}

x_labels <- c("500", "1k", "2k", "5k", "10k", "20k", "50k", "100k", "250k", "500k")

# Elbow: first bin where marginal R² gain drops below 5% of total gain (gene-rich region)
r2_gene   <- r2_all[, 2]
d1        <- diff(r2_gene)
total_gain <- sum(d1, na.rm = TRUE)
pct_gains  <- d1 / total_gain
elbow_idx  <- which(pct_gains < 0.05)[1]  # index into d1; corresponds to bin_sizes_c[elbow_idx+1]
elbow_x    <- bin_sizes_c[elbow_idx + 1]

par(mfrow = c(1, 2), mar = c(5, 4.5, 3, 1))

# R² panel
plot(NULL, xlim = range(bin_sizes_c), ylim = c(0, 1), log = "x",
     xlab = "Bin size (bp)", ylab = "R²",
     main = "R² vs bin size — three genomic regions",
     xaxt = "n")
axis(1, at = bin_sizes_c, labels = x_labels, cex.axis = 0.8)
abline(h = seq(0.2, 0.8, by = 0.2), col = "gray85", lty = 2, lwd = 0.7)
if (!is.na(elbow_x))
  abline(v = elbow_x, col = "gray40", lty = 3, lwd = 1.5)
for (j in seq_along(windows_c))
  lines(bin_sizes_c, r2_all[, j], type = "b", pch = 19, lwd = 2,
        col = windows_c[[j]]$col)
if (!is.na(elbow_x))
  text(elbow_x, 0.08, sprintf("elbow\n%s", x_labels[elbow_idx]),
       col = "gray40", cex = 0.75, adj = c(-0.1, 0))
legend("topleft",
       legend = sapply(windows_c, `[[`, "label"),
       col    = sapply(windows_c, `[[`, "col"),
       lwd = 2, pch = 19, bty = "n", cex = 0.85)

# Spearman panel
plot(NULL, xlim = range(bin_sizes_c), ylim = c(0, 1), log = "x",
     xlab = "Bin size (bp)", ylab = "Spearman r",
     main = "Spearman r vs bin size\nthree genomic regions",
     xaxt = "n")
axis(1, at = bin_sizes_c, labels = x_labels, cex.axis = 0.8)
abline(h = seq(0.2, 0.8, by = 0.2), col = "gray85", lty = 2, lwd = 0.7)
if (!is.na(elbow_x))
  abline(v = elbow_x, col = "gray40", lty = 3, lwd = 1.5)
for (j in seq_along(windows_c))
  lines(bin_sizes_c, sp_all[, j], type = "b", pch = 19, lwd = 2,
        col = windows_c[[j]]$col)

par(mfrow = c(1, 1))
```

All three regions show the same qualitative trend — R² and Spearman increase with bin size as aggregation smooths per-base noise.
The curves are vertically separated: the gene-rich region (green) sits consistently highest, the edge region (blue) is intermediate, and the centromere (red) has R² near zero regardless of bin size.

The Spearman panel reveals an important nuance: the centromere scores \~0.95 Spearman despite R² ≈ 0.
This is because Spearman only checks whether rank order is preserved — and in the centromere, the tiny fluctuations that exist happen to be consistent in rank.
But coverage is near-zero throughout, so there is no real signal to explain.
R² correctly reflects this; Spearman is misleading here.

```{r section-c-elbow, fig.width=6, fig.height=5}
bar_cols <- ifelse(pct_gains * 100 < 5, "firebrick", "#4DAF4A")
bar_labels <- x_labels[2:length(x_labels)]  # labels for transition points

barplot(pct_gains * 100,
        names.arg = bar_labels,
        col       = bar_cols,
        border    = NA,
        xlab      = "Bin size step (bp)",
        ylab      = "% of total R² gain",
        main      = "Marginal R² gain per bin size step\n(gene-rich region)",
        las       = 2,
        cex.names = 0.8)
abline(h = 5, col = "gray30", lty = 2, lwd = 1.5)
text(par("usr")[2], 5.5, "5% threshold", adj = 1, cex = 0.8, col = "gray30")
```

The bar chart shows what fraction of the total R² gain is contributed by each bin size step in the gene-rich region.
Green bars are steps that contribute meaningfully (≥5% of total gain); red bars fall below the threshold and represent diminishing returns.
The dashed line at 5% is the stopping criterion — once marginal gains drop below it, increasing bin size costs spatial resolution without meaningfully improving the model.

However, the statistical argument alone is not sufficient.
A 100 kb bin may satisfy the threshold, but it would swallow entire genes in a single bin — making it impossible to connect coverage signal back to any specific genomic feature.
Human genes are typically 10–100 kb long, and GC-rich regulatory elements such as CpG islands are only 0.5–2 kb.
At 5 kb bins, we get enough resolution to detect GC variation *within* genes, distinguish neighboring genes with different GC content, and partially capture CpG island signal — while still having enough reads per bin for stable estimates.
The optimal bin size is therefore bracketed by two constraints: the statistical elbow from above, and the scale of biologically meaningful features from below.
5 kb sits in the middle of that window.

------------------------------------------------------------------------

## Section d — Poisson Regression

```{r section-d-models}
# lm baseline
mod_lm   <- lm(reads ~ gc, data = df_main)

# Poisson GLM (log link, variance = mean)
mod_pois <- glm(reads ~ gc, data = df_main, family = poisson(link = "log"))

# Quasi-Poisson (variance = phi * mean — estimates dispersion directly)
mod_qpois <- glm(reads ~ gc, data = df_main, family = quasipoisson(link = "log"))

phi_hat <- summary(mod_qpois)$dispersion

# McFadden pseudo-R² for Poisson
pr2_pois  <- pseudo_r2(mod_pois)
pr2_qpois <- pseudo_r2(mod_qpois)

cat(sprintf("\n=== Dispersion parameter (phi) from quasi-Poisson: %.2f ===\n", phi_hat))
cat(sprintf("  (Lab 3 VMR was ~14.7 — consistent with overdispersion of this magnitude)\n\n"))

comparison_d <- data.frame(
  Model       = c("lm (Normal)", "glm Poisson", "glm quasi-Poisson"),
  R2_or_pR2   = round(c(summary(mod_lm)$r.squared, pr2_pois, pr2_qpois), 4),
  Metric      = c("R²", "McFadden pseudo-R²", "McFadden pseudo-R²"),
  AIC         = round(c(AIC(mod_lm), AIC(mod_pois), NA), 1),
  Dispersion  = c(NA, 1.0, round(phi_hat, 2)),
  stringsAsFactors = FALSE
)

cat("=== Model comparison table ===\n")
print(comparison_d, row.names = FALSE)
```

```{r section-d-plot, fig.width=10, fig.height=6}
gc_seq_d <- seq(min(df_main$gc), max(df_main$gc), length.out = 300)
pred_df_d <- data.frame(gc = gc_seq_d)

preds_lm    <- predict(mod_lm,    newdata = pred_df_d)
preds_pois  <- predict(mod_pois,  newdata = pred_df_d, type = "response")
preds_qpois <- predict(mod_qpois, newdata = pred_df_d, type = "response")

plot(df_main$gc, df_main$reads,
     col = rgb(0, 0, 0, 0.15), pch = 20, cex = 0.4,
     xlab = "GC Content", ylab = "Reads per bin",
     main = "lm vs Poisson vs quasi-Poisson",
     ylim = c(0, 400))
lines(gc_seq_d, preds_lm,    col = "steelblue", lwd = 2.5)
lines(gc_seq_d, preds_pois,  col = "red",       lwd = 2.5, lty = 2)
lines(gc_seq_d, preds_qpois, col = "darkgreen", lwd = 2.5, lty = 3)
legend("topleft",
       legend = c(
         sprintf("lm  (R²=%.4f)",             summary(mod_lm)$r.squared),
         sprintf("Poisson  (pR²=%.4f)",        pr2_pois),
         sprintf("quasi-Poisson  (φ=%.1f)",    phi_hat)
       ),
       col = c("steelblue", "red", "darkgreen"),
       lwd = 2.5, lty = c(1, 2, 3), bty = "n", cex = 1.0)
```

The quasi-Poisson model estimates a dispersion parameter φ ≈ `r round(phi_hat, 1)`, meaning the variance in read counts is roughly `r round(phi_hat, 1)`× what a true Poisson process would predict.
This directly corroborates the VMR ≈ 14.7 measured in lab 3 and the SD_ratio ≈ 2.25 from lab 4 — all three measurements reflect the same underlying overdispersion in this sequencing data.

------------------------------------------------------------------------

## Short Summary

We examined the GC-coverage relationship from four angles.

In (a), we found that G and C are nearly interchangeable predictors (Chargaff symmetry), and the interaction term adds negligible explanatory power.

In (b), the spatial analysis showed that the strength of the GC-coverage relationship varies considerably across chromosome 1, with R² ranging from near-zero in low-complexity regions (centromere, telomere) to above 0.8 in gene-rich regions.

In (c), larger bin sizes generally increased R², because aggregation smooths out per-base noise — though with diminishing returns beyond \~5000 bp.

In (d), Poisson and quasi-Poisson regression confirmed what we suspected since lab 3: the data are overdispersed (φ ≈ 14), and a quasi-Poisson model provides a more honest fit than either plain Poisson or linear regression.

------------------------------------------------------------------------

## Section e — AI Usage in This Lab

This lab was explicitly AI-first.
We used Claude Code throughout: to design the analysis plan, write the utility function library (`lab5_utils.R`), scaffold the Shiny app, debug graphical parameter issues, and refine the narrative.
Key decisions — model selection, which comparisons to make, what to highlight in the presentation — were made by us; Claude handled implementation speed.
The Shiny interactive explorer was designed collaboratively: we described the desired interaction (slider + live plot update) and Claude implemented it.

## External Sources Statement

We used AI (Claude Code) to:

1.  Design the overall analysis plan for this lab
2.  Write and debug `lab5_utils.R` utility functions
3.  Build the Shiny interactive application
4.  Fix plot layout and legend positioning
5.  Connect quasi-Poisson dispersion back to lab 3/4 findings
