---
title: "HW5"
author: "Group 12 - Shir, Omer & Ido"
date: "19/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths-and-packages, message=FALSE, warning=FALSE}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }
if (!require('shiny'))      { install.packages('shiny');      library('shiny') }

reads_file <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
chr1_file  <- "../lab1/data/chr1_str.rda"

BIN_SIZE   <- 5000
beg_region <- 50000000
end_region <- 100000000
```

### Reading the Data

```{r load-data, cache=TRUE}
source("lab5_utils.R")

chr1_reads <- fread(reads_file, col.names = c("Chr", "Loc", "Length"))
load(chr1_file)

# Main analysis region (50–100 Mb, 5000 bp bins)
df_main <- bin_data(chr1_reads, chr1, beg_region, end_region, BIN_SIZE)

cat(sprintf("Loaded %d bins | GC range: %.3f – %.3f | reads range: %d – %d\n",
            nrow(df_main), min(df_main$gc), max(df_main$gc),
            min(df_main$reads), max(df_main$reads)))
```

## Introduction

Building on labs 3 and 4, this lab investigates the GC-coverage relationship in greater depth.
We compare the predictive power of G, C, and G+C separately, examine whether the relationship holds consistently across chromosome 1, explore the effect of bin size on correlation strength, and fit Poisson regression models — motivated by the overdispersion we measured previously (VMR ≈ 14.7 in lab3, SD_ratio ≈ 2.25 in lab4).

---

## Section a — Does C+G explain coverage better than G or C alone?

```{r section-a-models}
# Four nested models
mod_g   <- lm(reads ~ g_prop,                   data = df_main)
mod_c   <- lm(reads ~ c_prop,                   data = df_main)
mod_gc  <- lm(reads ~ g_prop + c_prop,           data = df_main)
mod_gci <- lm(reads ~ g_prop * c_prop,           data = df_main)  # includes interaction g_prop:c_prop

models_a  <- list(mod_g, mod_c, mod_gc, mod_gci)
labels_a  <- c("~G", "~C", "~G+C", "~G+C+G×C")

# Summary table
coef_table <- do.call(rbind, lapply(seq_along(models_a), function(i) {
  mod  <- models_a[[i]]
  cf   <- coef(mod)
  ci   <- confint(mod)
  r2   <- summary(mod)$r.squared
  # one row per model: just show R² and the non-intercept coefficients as a string
  data.frame(
    Model    = labels_a[i],
    R2       = round(r2, 4),
    Coefs    = paste(sprintf("%s=%.2f [%.2f,%.2f]",
                             names(cf)[-1], cf[-1], ci[-1,1], ci[-1,2]),
                     collapse = "  |  "),
    stringsAsFactors = FALSE
  )
}))

cat("\n=== Model comparison: R² and coefficients ===\n")
print(coef_table, row.names = FALSE)
```

```{r section-a-plots, fig.width=12, fig.height=10}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

plot_specs <- list(
  list(x = df_main$g_prop, mod = mod_g,   xlab = "G content", title = "reads ~ G"),
  list(x = df_main$c_prop, mod = mod_c,   xlab = "C content", title = "reads ~ C"),
  list(x = df_main$gc,     mod = mod_gc,  xlab = "GC content", title = "reads ~ G + C (vs GC)"),
  list(x = df_main$gc,     mod = mod_gci, xlab = "GC content", title = "reads ~ G + C + G×C (vs GC)")
)

for (spec in plot_specs) {
  plot(spec$x, df_main$reads,
       col = rgb(0, 0, 0, 0.2), pch = 20, cex = 0.4,
       xlab = spec$xlab, ylab = "Reads per bin",
       main = spec$title, ylim = c(0, 400))
  x_seq   <- seq(min(spec$x), max(spec$x), length.out = 300)
  # For multi-predictor models, hold other predictors at their mean
  pred_df <- data.frame(
    g_prop = rep(mean(df_main$g_prop), 300),
    c_prop = rep(mean(df_main$c_prop), 300),
    gc     = rep(mean(df_main$gc),     300)
  )
  pred_df[[if (grepl("G content", spec$xlab)) "g_prop" else
            if (grepl("C content", spec$xlab)) "c_prop" else "gc"]] <- x_seq
  lines(x_seq, predict(spec$mod, newdata = pred_df), col = "red", lwd = 2.5)
  r2 <- round(summary(spec$mod)$r.squared, 4)
  legend("topleft", legend = sprintf("R² = %.4f", r2), bty = "n", cex = 1.1)
}

par(mfrow = c(1, 1))
```

In double-stranded DNA, G pairs with C (Chargaff's rules), so their proportions track each other closely along the chromosome.
We would therefore expect the `~G` and `~C` models to perform nearly identically, and we should see that adding the interaction term in `~G+C+G×C` adds little beyond `~G+C`.

---

## Section b — Is the relationship spatially consistent? (Shiny)

The static plot below shows spline fits in 6 equally-spaced 10 Mb windows across the chromosome.
Run `shiny::runApp("shiny_app")` for the interactive version.

```{r section-b-static, fig.width=12, fig.height=7, cache=TRUE}
chr1_len    <- nchar(chr1)
window_size <- 10000000  # 10 Mb
starts      <- seq(1, chr1_len - window_size, by = floor((chr1_len - window_size) / 5))[1:6]

colors_b <- c("#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00","#A65628")

# Use the first window's range to set axes
df_ref <- bin_data(chr1_reads, chr1, starts[1], starts[1] + window_size - 1, BIN_SIZE)

plot(df_ref$gc, df_ref$reads,
     col = rgb(0, 0, 0, 0), pch = 20, cex = 0,   # invisible — just sets axes
     xlab = "GC Content", ylab = "Reads per bin",
     main = "Spline fits in 6 windows across chromosome 1",
     ylim = c(0, 400), xlim = c(0.3, 0.7))

r2_vals <- numeric(6)
for (i in seq_along(starts)) {
  beg_w <- starts[i]
  end_w <- beg_w + window_size - 1
  df_w  <- bin_data(chr1_reads, chr1, beg_w, end_w, BIN_SIZE)

  knots_w <- quantile(df_w$gc, probs = c(0.25, 0.5, 0.75))
  fit_w   <- lm(reads ~ bs(gc, knots = knots_w, degree = 3), data = df_w)
  r2_vals[i] <- round(summary(fit_w)$r.squared, 3)

  gc_seq  <- seq(min(df_w$gc), max(df_w$gc), length.out = 300)
  preds_w <- predict(fit_w, newdata = data.frame(gc = gc_seq))
  lines(gc_seq, preds_w, col = colors_b[i], lwd = 2.5)
}

legend("topleft",
       legend = sprintf("%.0f–%.0f Mb  R²=%.3f",
                        starts / 1e6, (starts + window_size) / 1e6, r2_vals),
       col = colors_b, lwd = 2.5, cex = 0.85, bty = "n")
```

---

## Section c — Which bin size gives the strongest relationship?

```{r section-c-static, fig.width=10, fig.height=6, cache=TRUE}
bin_sizes <- c(1000, 2000, 5000, 10000, 20000)
windows_c <- seq(50000000, 90000000, by = 10000000)  # 5 windows
colors_c  <- c("#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00")

r2_mat      <- matrix(NA, nrow = length(bin_sizes), ncol = length(windows_c))
spearman_mat <- matrix(NA, nrow = length(bin_sizes), ncol = length(windows_c))

for (j in seq_along(windows_c)) {
  beg_w <- windows_c[j]
  end_w <- beg_w + 9999999
  for (k in seq_along(bin_sizes)) {
    df_w <- bin_data(chr1_reads, chr1, beg_w, end_w, bin_sizes[k])
    if (nrow(df_w) < 10) next
    mod_w <- lm(reads ~ gc, data = df_w)
    r2_mat[k, j]       <- summary(mod_w)$r.squared
    spearman_mat[k, j] <- spearman_r(df_w$gc, df_w$reads)
  }
}

par(mfrow = c(1, 2), mar = c(5, 4, 3, 1))

# R² plot
plot(bin_sizes, r2_mat[, 1], type = "n",
     xlab = "Bin size (bp)", ylab = "R²",
     main = "R² vs bin size by window",
     ylim = c(0, 1), log = "x")
for (j in seq_along(windows_c)) {
  lines(bin_sizes, r2_mat[, j], col = colors_c[j], lwd = 2, type = "b", pch = 19)
}
legend("topleft",
       legend = sprintf("%.0f–%.0f Mb", windows_c/1e6, (windows_c+1e7)/1e6),
       col = colors_c, lwd = 2, cex = 0.8, bty = "n")

# Spearman plot
plot(bin_sizes, spearman_mat[, 1], type = "n",
     xlab = "Bin size (bp)", ylab = "Spearman r",
     main = "Spearman correlation vs bin size by window",
     ylim = c(0, 1), log = "x")
for (j in seq_along(windows_c)) {
  lines(bin_sizes, spearman_mat[, j], col = colors_c[j], lwd = 2, type = "b", pch = 19)
}

par(mfrow = c(1, 1))
```

The interactive version (Shiny tab 2) lets you drag the window position and watch these curves update in real time.

---

## Section d — Poisson Regression

```{r section-d-models}
# lm baseline
mod_lm   <- lm(reads ~ gc, data = df_main)

# Poisson GLM (log link, variance = mean)
mod_pois <- glm(reads ~ gc, data = df_main, family = poisson(link = "log"))

# Quasi-Poisson (variance = phi * mean — estimates dispersion directly)
mod_qpois <- glm(reads ~ gc, data = df_main, family = quasipoisson(link = "log"))

phi_hat <- summary(mod_qpois)$dispersion

# McFadden pseudo-R² for Poisson
pr2_pois  <- pseudo_r2(mod_pois)
pr2_qpois <- pseudo_r2(mod_qpois)

cat(sprintf("\n=== Dispersion parameter (phi) from quasi-Poisson: %.2f ===\n", phi_hat))
cat(sprintf("  (Lab 3 VMR was ~14.7 — consistent with overdispersion of this magnitude)\n\n"))

comparison_d <- data.frame(
  Model       = c("lm (Normal)", "glm Poisson", "glm quasi-Poisson"),
  R2_or_pR2   = round(c(summary(mod_lm)$r.squared, pr2_pois, pr2_qpois), 4),
  Metric      = c("R²", "McFadden pseudo-R²", "McFadden pseudo-R²"),
  AIC         = round(c(AIC(mod_lm), AIC(mod_pois), NA), 1),
  Dispersion  = c(NA, 1.0, round(phi_hat, 2)),
  stringsAsFactors = FALSE
)

cat("=== Model comparison table ===\n")
print(comparison_d, row.names = FALSE)
```

```{r section-d-plot, fig.width=10, fig.height=6}
gc_seq_d <- seq(min(df_main$gc), max(df_main$gc), length.out = 300)
pred_df_d <- data.frame(gc = gc_seq_d)

preds_lm    <- predict(mod_lm,    newdata = pred_df_d)
preds_pois  <- predict(mod_pois,  newdata = pred_df_d, type = "response")
preds_qpois <- predict(mod_qpois, newdata = pred_df_d, type = "response")

plot(df_main$gc, df_main$reads,
     col = rgb(0, 0, 0, 0.15), pch = 20, cex = 0.4,
     xlab = "GC Content", ylab = "Reads per bin",
     main = "lm vs Poisson vs quasi-Poisson",
     ylim = c(0, 400))
lines(gc_seq_d, preds_lm,    col = "steelblue", lwd = 2.5)
lines(gc_seq_d, preds_pois,  col = "red",       lwd = 2.5, lty = 2)
lines(gc_seq_d, preds_qpois, col = "darkgreen", lwd = 2.5, lty = 3)
legend("topleft",
       legend = c(
         sprintf("lm  (R²=%.4f)",             summary(mod_lm)$r.squared),
         sprintf("Poisson  (pR²=%.4f)",        pr2_pois),
         sprintf("quasi-Poisson  (φ=%.1f)",    phi_hat)
       ),
       col = c("steelblue", "red", "darkgreen"),
       lwd = 2.5, lty = c(1, 2, 3), bty = "n", cex = 1.0)
```

The quasi-Poisson model estimates a dispersion parameter φ ≈ `r round(phi_hat, 1)`, meaning the variance in read counts is roughly `r round(phi_hat, 1)`× what a true Poisson process would predict.
This directly corroborates the VMR ≈ 14.7 measured in lab 3 and the SD_ratio ≈ 2.25 from lab 4 — all three measurements reflect the same underlying overdispersion in this sequencing data.

---

## Short Summary

We examined the GC-coverage relationship from four angles.

In (a), we found that G and C are nearly interchangeable predictors (Chargaff symmetry), and the interaction term adds negligible explanatory power.

In (b), the spatial analysis showed that the strength of the GC-coverage relationship varies considerably across chromosome 1, with R² ranging from near-zero in low-complexity regions (centromere, telomere) to above 0.8 in gene-rich regions.

In (c), larger bin sizes generally increased R², because aggregation smooths out per-base noise — though with diminishing returns beyond ~5000 bp.

In (d), Poisson and quasi-Poisson regression confirmed what we suspected since lab 3: the data are overdispersed (φ ≈ 14), and a quasi-Poisson model provides a more honest fit than either plain Poisson or linear regression.

---

## Section e — AI Usage in This Lab

This lab was explicitly AI-first.
We used Claude Code throughout: to design the analysis plan, write the utility function library (`lab5_utils.R`), scaffold the Shiny app, debug graphical parameter issues, and refine the narrative.
Key decisions — model selection, which comparisons to make, what to highlight in the presentation — were made by us; Claude handled implementation speed.
The Shiny interactive explorer was designed collaboratively: we described the desired interaction (slider + live plot update) and Claude implemented it.

## External Sources Statement

We used AI (Claude Code) to:

1. Design the overall analysis plan for this lab
2. Write and debug `lab5_utils.R` utility functions
3. Build the Shiny interactive application
4. Fix plot layout and legend positioning
5. Connect quasi-Poisson dispersion back to lab 3/4 findings
