---
title: "HW2"
author: "Group 12 - Shir, Omer & Ido"
date: "14/04/26"
output:
  pdf_document:
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths, packages & global vars}
if (!require('data.table')){
  install.packages('data.table')
  library('data.table')
}
library('tictoc')

if (!require('manipulate')){
  install.packages('manipulate')
  library('manipulate')
}


reads_file = "data/TCGA-13-0723-10B_lib2_all_chr1.forward"

beg_region <- 50000000
end_region <- 60000000

```

### Reading the Data:

```{r loading data}
chr1_reads = fread(reads_file, col.names = c("Chr", "Loc", "Length"))

```

## Introduction

This lab focuses on analyzing NGS fastq files, and specifically read coverage and depth.

## Part 1

### Question 1

The original, semi-naive implelmtatnion here is adapted from Yuval's second lecture, with additional logging for easier comparison.

```{r q1-naive-loop}
count_reads_naive <- function(reads, beg_region, end_region, log_every = 1e6) {
  N <- end_region - beg_region + 1
  read_line <- numeric(N)

  first_read <- min(which(reads$Loc >= beg_region))
  r <- first_read
  start_time <- proc.time()["elapsed"]

  while (reads$Loc[r] <= end_region) {
    read_line[reads$Loc[r] - beg_region + 1] <- read_line[reads$Loc[r] - beg_region + 1] + 1
    if ((r - first_read) %% log_every == 0) {
      elapsed <- round(proc.time()["elapsed"] - start_time, 1)
      cat("Progress: read", r, "| loc", reads$Loc[r], "| elapsed:", elapsed, "s\n")
      flush.console()
    }
    r <- r + 1
  }
  read_line
}
```

```{r q1-run}
tic()
counts_naive <- count_reads_naive(chr1_reads, beg_region, end_region)
t_naive <- toc(quiet = TRUE)

message("Total fragments in region: ", sum(counts_naive))
message("Time taken: ", round(t_naive$toc - t_naive$tic, 2), "s")
```

### Question 2

To find a better implementaiton, we looked at several options suggested in StackOverflow (links below), before finally settling on this one that uses `tabulate`.

We rely mostly on the assumption that the location list is already sorted (if it wasn't, it would have been an O(nlogn) task instead).
The reason for our expectation for this to be faster is empirical- we tried several options mentioned in StackOverflow, and this one seemed to do the trick.

```{r q2-table}
source("../utils.R")
```

```{r q2-run}
tic()
counts_table <- count_reads_table(chr1_reads, beg_region, end_region)
t_table <- toc(quiet = TRUE)

message("Total fragments in region: ", sum(counts_table))
message("Time taken: ", round(t_table$toc - t_table$tic, 2), "s")
```

### Question 3

To compare the functions, we create a benchmark test.
the tests randomly selects a 1M window for 100 times.
it run each function over that window, and stores the result.
Here is the benchmark:

```{r q3-benchmark}
chr1_max_loc <- 247199205
window_size  <- 1e6
n_runs       <- 100

times_naive  <- numeric(n_runs)
times_tabulate <- numeric(n_runs)

set.seed(42)
starts <- sample(1:(chr1_max_loc - window_size), n_runs)

for (i in 1:n_runs) {
  beg <- starts[i]
  end <- beg + window_size - 1

  t0 <- proc.time()["elapsed"]
  suppressMessages(count_reads_naive(chr1_reads, beg, end))
  times_naive[i] <- proc.time()["elapsed"] - t0

  t0 <- proc.time()["elapsed"]
  suppressMessages(count_reads_table(chr1_reads, beg, end))
  times_tabulate[i] <- proc.time()["elapsed"] - t0

  if (i %% 10 == 0) { cat("Run", i, "/", n_runs, "\n"); flush.console() }
}
```

And the results:

```{r q3-plot}
x_max <- 0.08
breaks <- seq(0, x_max, length.out = 50)
h_naive    <- hist(pmin(times_naive, x_max),    breaks = breaks, plot = FALSE)
h_tabulate <- hist(pmin(times_tabulate, x_max), breaks = breaks, plot = FALSE)

y_max <- max(h_naive$counts, h_tabulate$counts) * 1.1
plot(0, type = "n", xlim = c(0, 0.08), ylim = c(-y_max, y_max),
     xlab = "Time (seconds)", ylab = "Count",
     main = "Runtime Distribution: Sorted Walk (down) vs. Tabulate (up)\n(100 random 1M-bp windows, chr1)",
     yaxt = "n")

yticks <- pretty(c(0, y_max))
yticks <- yticks[yticks > 0]  # exclude 0 to avoid duplication at center
axis(2, at = c(-yticks, 0, yticks), labels = c(yticks, 0, yticks))

rect(h_tabulate$breaks[-length(h_tabulate$breaks)], 0,
     h_tabulate$breaks[-1], h_tabulate$counts,
     col = "#56B4E9", border = "white")
rect(h_naive$breaks[-length(h_naive$breaks)], 0,
     h_naive$breaks[-1], -h_naive$counts,
     col = "#E69F00", border = "white")

abline(h = 0, lwd = 1)
abline(v = mean(times_naive),    col = "#E69F00", lwd = 2, lty = 2)
abline(v = mean(times_tabulate), col = "#56B4E9", lwd = 2, lty = 2)

legend("topright",
       legend = c(sprintf("Tabulate (mean=%.3fs)", mean(times_tabulate)),
                  sprintf("Sorted Walk (mean=%.3fs)", mean(times_naive))),
       fill = c("#56B4E9", "#E69F00"))
```

We can clearly see that the tabulate algorithm is superior to the naive sorted-walk algorithm seen in the lecture.

### Question 4

As we've done in the lecture, creating a histogram of the marginal coverage distribution:

```{r q4-coverage-hist}
covered <- counts_table[counts_table > 0]
max_cov <- min(max(covered), 50)

read_hist <- hist(pmin(covered, max_cov),
                  breaks = seq(0.5, max_cov + 0.5, by = 1),
                  plot = FALSE)

# log="y" in plot.histogram drops bars — draw manually
plot(read_hist$mids, read_hist$counts,
     type = "n", log = "y",
     xlim = range(read_hist$breaks),
     ylim = c(1, max(read_hist$counts) * 2),
     xlab = "Coverage (reads per base)",
     ylab = "Number of bases (log scale)",
     main = sprintf("Base-level Coverage Distribution (covered bases only)\n(chr1, %dM-%dM bp)",
                    beg_region/1e6, end_region/1e6),
     xaxt = "n")

rect(read_hist$breaks[-length(read_hist$breaks)], 1,
     read_hist$breaks[-1], read_hist$counts,
     col = "#56B4E9", border = "white")

axis(1, at = 1:max_cov)

text(read_hist$mids, read_hist$counts, read_hist$counts, pos = 3, cex = 0.9)
```

Note that we log-scaled the graph so the 1st bin will fit.
We can see the same tailed distribution we saw in class.

## Part 2

### Question 5.1 & 5.2:

```{r}
# Define the bin size (must be a multiple of 10,000)
BIN_SIZE <- 10000 

# Function to summarize base-level coverage into genomic bins
bin_coverage <- function(base_counts, bin_size) {
  n_bins <- floor(length(base_counts) / bin_size)
  counts_trimmed <- base_counts[1:(n_bins * bin_size)]
  
  # Reshape into matrix and calculate column sums for efficiency
  bin_matrix <- matrix(counts_trimmed, nrow = bin_size)
  bin_sums <- colSums(bin_matrix)
  
  return(bin_sums)
}
```

```{r fig.width=10, fig.height=8}
# Compute full-chromosome coverage using tabulate (no region filter needed)
full_chr_counts <- tabulate(chr1_reads$Loc)
binned_data <- bin_coverage(full_chr_counts, BIN_SIZE)
bin_positions_mb <- (seq_along(binned_data) * BIN_SIZE) / 1e6

zoom_start <- 120
zoom_end <- 150
y_cap <- 1500

par(mfrow = c(2, 1))

# 5.1: Full chromosome coverage plot (capped y-axis, outliers marked)
plot(bin_positions_mb, pmin(binned_data, y_cap), 
     type = "l", col = "steelblue",
     xlab = "Position on Chromosome 1 (Mb)", 
     ylab = "Reads per bin",
     main = "Full Chromosome Coverage",
     ylim = c(0, y_cap))
abline(h = mean(binned_data), col = "red", lty = 2)
abline(v = c(zoom_start, zoom_end), col = "black", lty = 1, lwd = 1.5)
# mark outlier bins above cap
outliers <- which(binned_data > y_cap)
points(bin_positions_mb[outliers], rep(y_cap, length(outliers)),
       col = "red", pch = 17, cex = 0.7)
legend("topright",
       legend = c(sprintf("Mean (%.0f)", mean(binned_data)), "Zoom region",
                  sprintf("Outliers (>%d)", y_cap)),
       col = c("red", "black", "red"), lty = c(2, 1, NA),
       pch = c(NA, NA, 17), lwd = c(1, 1.5, NA), cex = 0.8)

# 5.2: Zoom-in plot (20Mb window)
plot(bin_positions_mb, binned_data, 
     type = "l", col = "darkorange",
     xlim = c(zoom_start, zoom_end),
     ylim = c(0, y_cap),
     xlab = "Position on Chromosome 1 (Mb)", 
     ylab = "Reads per bin",
     main = sprintf("Zoom-in (%dMb - %dMb)", zoom_start, zoom_end))
points(bin_positions_mb, binned_data, pch = 20, cex = 0.5, col = "darkorange")


```

## Discussion

Based on the plots, we can spot several anomalies.
In the first plot of the full chromosome, the area around 120-130Mb shows a clear gap with coverage of almost 0, which is related to the centeromer we mentioned in class.

Right after the this part (\~130-145Mb), we see extreme outlier spikes, likely caused by reads mapping to duplicated segments.
Additionally, the beginning of the chromosome (0-30Mb) also shows high coverage above the mean, which might indicate that several segments there were also copied multiple times.

### Question 6

```{r}
# Filter out extreme outlier bins before fitting
binned_no_outliers <- binned_data[binned_data <= y_cap]

# Calculate parameters for the Normal distribution
mu <- mean(binned_no_outliers)
sigma <- sd(binned_no_outliers)

# Pre-calculate histogram and density curve to set proper axis limits
h_data <- hist(binned_no_outliers, breaks = 30, plot = FALSE)
x_vals <- seq(min(binned_no_outliers), max(binned_no_outliers), length = 100)
y_vals <- dnorm(x_vals, mean = mu, sd = sigma)

# Plot the histogram using freq = FALSE to correctly show density
# This fixes the warning and ensures bars align with the normal curve
plot(h_data, freq = FALSE,
     col = "lightblue", border = "white",
     xlab = "Reads per Bin",
     ylab = "Density",
     main = sprintf("Marginal Distribution vs. Normal Fit\n(\u03bc=%.0f, \u03c3=%.0f)", mu, sigma),
     ylim = c(0, max(h_data$density, y_vals) * 1.3))

# Overlay the Normal density curve
lines(x_vals, y_vals, col = "darkred", lwd = 2)

# Add legend to the top-left to avoid overlapping with the distribution peak
legend("topleft", legend = c("Observed Data", "Normal Fit"), 
       fill = c("lightblue", NA), border = c("black", NA),
       lty = c(NA, 1), col = c(NA, "darkred"), lwd = c(NA, 2),
       bty = "n")



```

The normal distribution does not seem to fit well at all - we could argue that the middle is somewhat normal, but even that doesn't seem correct, as it's not really bell-shaped.
the main differences are in the heavy left tail of the distribution, and in the missing right tail of it.
also the middle part looks more uniform then we'd expect.

## Short summary

In the first part, we experimented with optimizing the vector-reading function we saw in class.
using the built-in , C-based functions we managed to do that pretty well.

In the second part, we mapped the coverage across the entire chromosome using the NGS data.
we looked for interesting and irregular regions, and tried to compare the distribution to a normal one.

## External sources statement

We read a few articles in StackOverflow to find reccomendations for optimizing the R function.
Specifically relyed on this [article](https://stackoverflow.com/questions/17223308/fastest-way-to-count-occurrences-of-each-unique-element), and this specific [comment](https://stackoverflow.com/questions/17223308/fastest-way-to-count-occurrences-of-each-unique-element#17223496).

We also used AI (Claude) to:

1.  help us correctly use the tabulate syntax in question 2
2.  To create the back-to-back histogram using rectangles
3.  help us debug missing rectangle bug in question 4 after using log transformation.
4.  help us implement an efficient binning strategy using matrix operations and colSums.
5.  Assist with the multi-panel plot layout and fine-tuning the legend positioning to ensure all data remained visible.
