---
title: "HW2"
author: "Group 12 - Shir, Omer & Ido"
date: "14/04/26"
output:
  pdf_document:
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths, packages & global vars}
if (!require('data.table')){
  install.packages('data.table')
  library('data.table')
}
library('tictoc')

if (!require('manipulate')){
  install.packages('manipulate')
  library('manipulate')
}


reads_file = "data/TCGA-13-0723-10B_lib2_all_chr1.forward"

beg_region <- 50000000
end_region <- 60000000

```
### Reading the Data:

```{r loading data}
chr1_reads = fread(reads_file, col.names = c("Chr", "Loc", "Length"))

```

## Introduction
This lab focuses on analyzing NGS fastq files, and specifically read coverage and depth. 

## Part 1
### Question 1

The original, semi-naive implelmtatnion here is adapted from Yuval's second lecture, with additional logging for easier comparison.

```{r q1-naive-loop}
count_reads_naive <- function(reads, beg_region, end_region, log_every = 1e6) {
  N <- end_region - beg_region + 1
  read_line <- numeric(N)

  first_read <- min(which(reads$Loc >= beg_region))
  r <- first_read
  start_time <- proc.time()["elapsed"]

  while (reads$Loc[r] <= end_region) {
    read_line[reads$Loc[r] - beg_region + 1] <- read_line[reads$Loc[r] - beg_region + 1] + 1
    if ((r - first_read) %% log_every == 0) {
      elapsed <- round(proc.time()["elapsed"] - start_time, 1)
      cat("Progress: read", r, "| loc", reads$Loc[r], "| elapsed:", elapsed, "s\n")
      flush.console()
    }
    r <- r + 1
  }
  read_line
}
```

```{r q1-run}
tic()
counts_naive <- count_reads_naive(chr1_reads, beg_region, end_region)
t_naive <- toc(quiet = TRUE)

message("Total fragments in region: ", sum(counts_naive))
message("Time taken: ", round(t_naive$toc - t_naive$tic, 2), "s")
```

### Question 2

To find a better implementaiton, we looked at several options suggested in StackOverflow (links below), before finally settling on this one that uses `tabulate`. 

We rely mostly on the assumption that the location list is already sorted (if it wasn't, it would have been an O(nlogn) task instead). The reason for our expectation for this to be faster is empirical-  we tried several options mentioned in StackOverflow, and this one seemed to do the trick.

```{r q2-table}
count_reads_table <- function(reads, beg_region, end_region) {
  t0 <- proc.time()["elapsed"]

  # keep only reads that fall within the region
  locs <- reads$Loc[reads$Loc >= beg_region & reads$Loc <= end_region]
  message("Filtering done: ", length(locs), " reads in region | ",
          round(proc.time()["elapsed"] - t0, 2), "s")

  # tabulate uses 1-based indexing, so shift locs to be relative to beg_region
  counts <- tabulate(locs - beg_region + 1, nbins = end_region - beg_region + 1)
  result <- counts
  message("Done | total: ", round(proc.time()["elapsed"] - t0, 2), "s")

  result
}
```

```{r q2-run}
tic()
counts_table <- count_reads_table(chr1_reads, beg_region, end_region)
t_table <- toc(quiet = TRUE)

message("Total fragments in region: ", sum(counts_table))
message("Time taken: ", round(t_table$toc - t_table$tic, 2), "s")
```
### Question 3
To compare the functions, we create a benchmark test. the tests randomly selects a 1M window for 100 times. it run each function over that window, and stores the result. Here is the benchmark:

```{r q3-benchmark}
chr1_max_loc <- 247199205
window_size  <- 1e6
n_runs       <- 100

times_naive  <- numeric(n_runs)
times_tabulate <- numeric(n_runs)

set.seed(42)
starts <- sample(1:(chr1_max_loc - window_size), n_runs)

for (i in 1:n_runs) {
  beg <- starts[i]
  end <- beg + window_size - 1

  t0 <- proc.time()["elapsed"]
  suppressMessages(count_reads_naive(chr1_reads, beg, end))
  times_naive[i] <- proc.time()["elapsed"] - t0

  t0 <- proc.time()["elapsed"]
  suppressMessages(count_reads_table(chr1_reads, beg, end))
  times_tabulate[i] <- proc.time()["elapsed"] - t0

  if (i %% 10 == 0) { cat("Run", i, "/", n_runs, "\n"); flush.console() }
}
```


And the results:

```{r q3-plot}
x_max <- 0.08
breaks <- seq(0, x_max, length.out = 50)
h_naive    <- hist(pmin(times_naive, x_max),    breaks = breaks, plot = FALSE)
h_tabulate <- hist(pmin(times_tabulate, x_max), breaks = breaks, plot = FALSE)

y_max <- max(h_naive$counts, h_tabulate$counts) * 1.1
plot(0, type = "n", xlim = c(0, 0.08), ylim = c(-y_max, y_max),
     xlab = "Time (seconds)", ylab = "Count",
     main = "Runtime Distribution: Sorted Walk (down) vs. Tabulate (up)\n(100 random 1M-bp windows, chr1)",
     yaxt = "n")

yticks <- pretty(c(0, y_max))
yticks <- yticks[yticks > 0]  # exclude 0 to avoid duplication at center
axis(2, at = c(-yticks, 0, yticks), labels = c(yticks, 0, yticks))

rect(h_tabulate$breaks[-length(h_tabulate$breaks)], 0,
     h_tabulate$breaks[-1], h_tabulate$counts,
     col = "#56B4E9", border = "white")
rect(h_naive$breaks[-length(h_naive$breaks)], 0,
     h_naive$breaks[-1], -h_naive$counts,
     col = "#E69F00", border = "white")

abline(h = 0, lwd = 1)
abline(v = mean(times_naive),    col = "#E69F00", lwd = 2, lty = 2)
abline(v = mean(times_tabulate), col = "#56B4E9", lwd = 2, lty = 2)

legend("topright",
       legend = c(sprintf("Tabulate (mean=%.3fs)", mean(times_tabulate)),
                  sprintf("Sorted Walk (mean=%.3fs)", mean(times_naive))),
       fill = c("#56B4E9", "#E69F00"))
```
We can clearly see that the tabulate algorithm is superior to the naive sorted-walk algorithm seen in the lecture.


### Question 4
As we've done in the lecture, creating a histogram of the marginal coverage distribution:

```{r q4-coverage-hist}
covered <- counts_table[counts_table > 0]
max_cov <- min(max(covered), 50)

read_hist <- hist(pmin(covered, max_cov),
                  breaks = seq(0.5, max_cov + 0.5, by = 1),
                  plot = FALSE)

# log="y" in plot.histogram drops bars — draw manually
plot(read_hist$mids, read_hist$counts,
     type = "n", log = "y",
     xlim = range(read_hist$breaks),
     ylim = c(1, max(read_hist$counts) * 2),
     xlab = "Coverage (reads per base)",
     ylab = "Number of bases (log scale)",
     main = sprintf("Base-level Coverage Distribution (covered bases only)\n(chr1, %dM-%dM bp)",
                    beg_region/1e6, end_region/1e6),
     xaxt = "n")

rect(read_hist$breaks[-length(read_hist$breaks)], 1,
     read_hist$breaks[-1], read_hist$counts,
     col = "#56B4E9", border = "white")

axis(1, at = 1:max_cov)

text(read_hist$mids, read_hist$counts, read_hist$counts, pos = 3, cex = 0.9)
```

We can see the same tailed distribution we saw in class.

## Part 2
### Question 5.1 & 5.2:
```{r}
# --- Question 5.1: Coverage across the analyzed region ---
# Define the bin size (must be a multiple of 10,000)
BIN_SIZE <- 10000 

# Function to summarize base-level coverage into genomic bins
bin_coverage <- function(base_counts, bin_size) {
  n_bins <- floor(length(base_counts) / bin_size)
  counts_trimmed <- base_counts[1:(n_bins * bin_size)]
  
  # Reshape into matrix and calculate column sums for efficiency
  bin_matrix <- matrix(counts_trimmed, nrow = bin_size)
  bin_sums <- colSums(bin_matrix)
  
  return(bin_sums)
}
```

```{r}
# Set up a 1x2 plotting area to see both graphs side-by-side
par(mfrow = c(1, 2))

# 5.1: Full region coverage plot
plot(bin_positions_mb, binned_data, 
     type = "l", col = "steelblue",
     xlab = "Position on Chromosome 1 (Mb)", 
     ylab = "Reads per bin",
     main = "Full Region Coverage")
abline(h = mean(binned_data), col = "red", lty = 2)

# 5.2: Zoom-in plot (2Mb window)
# Adjust these values based on your region of interest
zoom_start <- 52
zoom_end <- 54
plot(bin_positions_mb, binned_data, 
     type = "l", col = "darkorange",
     xlim = c(zoom_start, zoom_end),
     xlab = "Position on Chromosome 1 (Mb)", 
     ylab = "Reads per bin",
     main = "Zoom-in (52Mb - 54Mb)")
points(bin_positions_mb, binned_data, pch = 20, cex = 0.5, col = "darkorange")


```
## Discussion
Based on the plots, we identify a clear genomic anomaly. In the Full Region Coverage plot, the area between 53Mb and 56Mb shows a significant and persistent increase in read depth, staying above the global mean. This elevated baseline suggests a potential Copy Number Gain in this segment. Conversely, the region from 50Mb to 52Mb represents a lower, likely diploid, background coverage.



### Question 6
```{r}
# Calculate parameters for the Normal distribution
mu <- mean(binned_data)
sigma <- sd(binned_data)

# Pre-calculate histogram and density curve to set proper axis limits
h_data <- hist(binned_data, breaks = 30, plot = FALSE)
x_vals <- seq(min(binned_data), max(binned_data), length = 100)
y_vals <- dnorm(x_vals, mean = mu, sd = sigma)

# Plot the histogram using freq = FALSE to correctly show density
# This fixes the warning and ensures bars align with the normal curve
plot(h_data, freq = FALSE,
     col = "lightblue", border = "white",
     xlab = "Reads per Bin",
     ylab = "Density",
     main = "Marginal Distribution vs. Normal Fit",
     ylim = c(0, max(h_data$density, y_vals) * 1.3))

# Overlay the Normal density curve
lines(x_vals, y_vals, col = "darkred", lwd = 2)

# Add legend to the top-left to avoid overlapping with the distribution peak
legend("topleft", legend = c("Observed Data", "Normal Fit"), 
       fill = c("lightblue", NA), border = c("black", NA),
       lty = c(NA, 1), col = c(NA, "darkred"), lwd = c(NA, 2),
       bty = "n")



```
The Normal distribution provides a good representation for the center of the data, as expected from the Central Limit Theorem for large bins. However, heavy tails can be observed, indicating a slight deviation from the ideal model. These deviations stem from the regions we identified with high copy numbers, which produce outlier coverage values that exceed the normal diploid average.



## Short summary

In the first part, we experimented with optimizing the vector-reading function we saw in class. using the built-in , C-based functions we managed to do that pretty well. 

In the second part,  we utilized genomic binning to identify a potential copy number gain between 53Mb and 56Mb and verified the data's fit to a Normal distribution.

## External sources statement
We read a few articles in StackOverflow to find reccomendations for optimizing the R function.
Specifically relyed on this [article](https://stackoverflow.com/questions/17223308/fastest-way-to-count-occurrences-of-each-unique-element), and this specific [comment](https://stackoverflow.com/questions/17223308/fastest-way-to-count-occurrences-of-each-unique-element#17223496).

We also used AI (Claude) to:

1. help us correctly use the tabulate syntax in question 2
2. To create the back-to-back histogram using rectangles
3. help us debug missing rectangle bug in question 4 after using log transformation.
4. help us implement an efficient binning strategy using matrix operations and colSums.
5. Assist with the multi-panel plot layout and fine-tuning the legend positioning to ensure all data remained visible.
