---
title: "HW4"
author: "Group 12 - Shir, Omer & Ido"
date: "10/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths, packages & global vars}

```

### Reading the Data:

```{r loading data}

```

## Introduction
In this lab we are fitting a non-parametric (homoscedastic) spline regression,
to estimate the reads-GC content relationship. Our assumption is that the the relationship is continuous and therefore a spline regression is a valid tool for modelling it. Once we have this function, we can "remove" it from the actual reads data and analyze the connection between reads and mutation in the chromosome.


## Part 1


## Short summary


## External sources statement

We also used AI (Claude) to:
