---
title: "HW3"
author: "Group 12 - Shir, Omer & Ido"
date: "03/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths, packages & global vars}
if (!require('data.table'))  { install.packages('data.table');  library('data.table') }
if (!require('dplyr'))       { install.packages('dplyr');       library('dplyr') }
if (!require('kableExtra'))  { install.packages('kableExtra');  library('kableExtra') }

reads_file  <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
beg_region  <- 50000000
end_region  <- 60000000
```

### Reading the Data:

```{r loading data}
chr1_reads <- fread(reads_file, col.names = c("Chr", "Loc", "Length"))
```

## Introduction

In this lab we model base-level read coverage using the Poisson distribution and assess how well it fits the data.

## Part 1

### Question 1

**Computing λ — mean coverage per base:**

```{r compute-lambda}
source("../utils.R")

N      <- end_region - beg_region + 1
cov    <- count_reads_table(chr1_reads, beg_region, end_region)
lambda <- mean(cov)
cat(sprintf("lambda = %.4f\n", lambda))
```

**Expected Poisson distribution and comparison:**

```{r poisson-comparison, fig.width=7, fig.height=5}
# observed proportions
obs_counts <- tabulate(cov, nbins = max(cov) + 1)   # index k+1 = count of k
obs_prop   <- obs_counts / N

# expected Poisson probabilities
k_vals   <- 0:max(cov)
exp_prop <- dpois(k_vals, lambda)

# --- Graph ---
x_max    <- 6
k_show   <- k_vals[k_vals <= x_max]
obs_show <- pmax(obs_prop[k_show + 1], 1e-10)   # avoid log(0)
exp_show <- pmax(exp_prop[k_show + 1], 1e-10)

y_min <- 10^floor(log10(min(obs_show[obs_show > 1e-9], exp_show[exp_show > 1e-9])))

plot(k_show, obs_show,
     type = "h", lwd = 8, col = "#56B4E9", log = "y",
     xlab = "Number of reads starting at a base (k)",
     ylab = "Proportion of bases (log scale)",
     main = sprintf("Observed vs. Poisson(%s) coverage distribution\n(chr1, %dMb\u2013%dMb)",
                    signif(lambda, 2), beg_region/1e6, end_region/1e6),
     xlim = c(-0.5, x_max + 0.5),
     ylim = c(y_min, 1),
     xaxt = "n")
axis(1, at = k_show)
lines(k_show + 0.2, exp_show, type = "h", lwd = 3, col = "#E69F00")
legend("topright",
       legend = c("Observed", sprintf("Poisson(\u03bb = %s)", signif(lambda, 2))),
       col    = c("#56B4E9", "#E69F00"),
       lwd    = c(8, 3),
       bty    = "n")
```

The graph shows the expected poisson reads histogram compared to the observed histogram.
Before we used log-scale it was hard to distinguish exactly since the distribution with lambda =0.036 is skewed.
We can see that Poisson distribution does not really match the actual data, which is less left-skewed and more right-skewed.

```{r poisson-table}


obs_vals <- c(sapply(0:5, function(k) sum(cov == k) / N), sum(cov > 5) / N)
exp_vals <- c(dpois(0:5, lambda), ppois(5, lambda, lower.tail = FALSE))

# fixed decimal places: enough to show variation (all values are small proportions)
fmt <- function(x) formatC(x, format = "f", digits = 7)

data.frame(k = c(as.character(0:5), ">5"),
           Observed = fmt(obs_vals),
           Expected = fmt(exp_vals)) %>%
  knitr::kable(align = "c",
               caption = sprintf("Observed vs. Expected Poisson proportions (λ = %s)",
                                 signif(lambda, 2))) %>%
  kable_styling(bootstrap_options = c("bordered", "striped"), full_width = FALSE) %>%
  column_spec(1:3, border_left = TRUE, border_right = TRUE)
```

It's harder for us to notice the difference in the table compared to the graph, but from the graph it's clear that the Poisson distribution is not a very good fit.

### Question 2

```{r q2-setup}
BIN_SIZE <- 5000

bin_counts <- sapply(seq(1, N - BIN_SIZE + 1, by = BIN_SIZE), function(i) {
  sum(cov[i:(i + BIN_SIZE - 1)])
}) # for each bin, we apply for the sum function over cov vector.

lambda_bin <- lambda * BIN_SIZE  # expected mean under Poisson
```

**a. Robust estimators vs. Poisson prediction:**

```{r q2a}
med    <- median(bin_counts)
iqr    <- IQR(bin_counts)
# For Poisson, IQR ≈ 1.35 * sqrt(lambda) (normal approximation)
exp_iqr <- qpois(0.75, lambda_bin) - qpois(0.25, lambda_bin)

cat(sprintf("Median:       %.2f  (Poisson mean: %.2f)\n", med, lambda_bin))
cat(sprintf("IQR:          %.2f  (Poisson expected IQR ≈ %.2f)\n", iqr, exp_iqr))
```

As we've seen in class, we use the median as a robust estimate of the center instead of the average.
For spread, we use IQR as a robust estimator.

We can see that the center is pretty close to the observed distribution with a small bias.
The spread, however, is almost 5 times bigger than we expected under the Poisson model.

**b. Histogram vs. Poisson:**

```{r q2b, fig.width=7, fig.height=5}
x_max_b   <- quantile(bin_counts, 0.99)
bins_trim <- bin_counts[bin_counts <= x_max_b]

h <- hist(bins_trim, breaks = 30, plot = FALSE)

# smooth Poisson curve: dpois at every integer in range
k_range   <- 0:max(bin_counts)
pois_dens <- dpois(k_range, lambda_bin)

plot(h, freq = FALSE,
     col = "#56B4E9", border = "white",
     xlim = c(min(h$breaks), max(bin_counts)),
     ylim = c(0, max(h$density, pois_dens) * 1.1),
     xlab = sprintf("Reads per bin (%d bp)", BIN_SIZE),
     ylab = "Density",
     main = sprintf("Fragment counts per bin vs. Poisson(\u03bb = %s)\n(chr1, %dMb\u2013%dMb)",
                    signif(lambda_bin, 3), beg_region/1e6, end_region/1e6))
lines(k_range, pois_dens, col = "#E69F00", lwd = 2)

legend("topright",
       legend = c("Observed", sprintf("Poisson(\u03bb = %s)", signif(lambda_bin, 3))),
       fill   = c("#56B4E9", NA), border = c("black", NA),
       lty    = c(NA, 1), col = c(NA, "#E69F00"), lwd = c(NA, 2),
       bty    = "n")
```

In the lambda value we got, 177, the expected distribution looks almost identical to a normal distribution.
As we mentioned in a, the center is aligned with the expectation, but the observed data spread is much higher, indicating once again that the Poisson model is not a good fit.

**c. Variance-to-Mean Ratio (VMR):**

```{r q2c}
vmr <- var(bin_counts) / mean(bin_counts)
cat(sprintf("Mean:     %.4f\n", mean(bin_counts)))
cat(sprintf("Variance: %.4f\n", var(bin_counts)))
cat(sprintf("VMR:      %.4f  (Poisson expectation: 1)\n", vmr))
```

For a Poisson distribution, VMR = 1 by definition (variance equals mean).
The VMR we actually got, 14.6974, again indicates this model is not a good choice for our specific case.
Perhaps one of our assumptions - uniformity or independence is incorrect.

### Question 3

```{r q3-setup}
load("../lab1/data/chr1_str.rda")

seq_region <- substr(chr1, 50000000, 100000000)
split_seq  <- strsplit(seq_region, "")[[1]]

# GC count per bin — upper bound prevents last bin from going out of bounds
gc_binned <- sapply(seq(1, length(split_seq) - BIN_SIZE + 1, by = BIN_SIZE), function(i) {
  sum(split_seq[i:(i + BIN_SIZE - 1)] %in% c("C", "G", "c", "g"))
})

min_len <- min(length(bin_counts), length(gc_binned))
cov_sub <- bin_counts[1:min_len]
gc_sub  <- gc_binned[1:min_len]
```

**a. Calculate correlation**

```{r q3a}
cor_val <- cor(gc_sub, cov_sub)
cat(sprintf("Correlation between GC content and coverage: %.4f\n", cor_val))
```

The calculated correlation coefficient is 0.8207, showing a strong positive linear relationship between the GC content and the observed read coverage per bin.
Unlike the findings by Dohm which typically show reduced coverage at extreme GC levels, our region shows a strong positive relationship where higher GC content yields higher read coverage.

```{r q3b}
plot(gc_sub, cov_sub,
     xlab = "GC count per bin",
     ylab = "Number of reads (coverage)",
     main = "GC Content vs. Binned Coverage",
     pch = 20, col = rgb(0, 0.3, 0.7, 0.4), cex = 0.8)
abline(lm(cov_sub ~ gc_sub), col = "red", lwd = 2)
```

From the plot we can observe that there is a strong positive linear relation between the GC content and the read coverage.
Unlike the finding in the Dohm article, where coverage drops at extreme GC levels- our data shows a steady increase in coverage as GC content rises.

**c. Relationship to Part 1** The strong positive correlation between GC content and coverage explains the large spread and high variance-mean ratio observed in Part 1.
The assumption of a uniform Poisson distribution doesn't hold since read coverage is not random or even across the chromosome.
Instead, it is driven, to some extent, by local GC content.

## Short summary

In the first part we tried to asses the Poisson model fit for our data, and saw it has significant inconsistencies with the observed distribution.

In the second part, we used genomic binning to looked at the correlation between GC content and read coverage.

## External sources statement

We also used AI (Claude) to:

1.  move the function from lab2 to a utils file
2.  edit the graph legend in q1.
3.  use `knitr::kable` for table creation in q1.
4.  correctly use `qpois` to calculate the expected IQR
