---
title: "HW6"
author: "Group 12 - Shir, Omer & Ido"
date: "26/05/26"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["fvextra"]
  html_document:
    code_folding: show
header-includes:
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options:
  markdown:
    wrap: sentence
---

### Group:

Group members (Name, ID, email)

-   Ido Ravid, 207198177, ido.ravid\@mail.huji.ac.il

-   Omer Sutovsky, 318735107 ,omer.sutovsky\@mail.huji.ac.il

-   Shir Tsarfaty, 207718008, shir.tsarfaty\@mail.huji.ac.il

### Paths and Data

```{r paths_and_data}
if (!require('data.table')) { install.packages('data.table'); library('data.table') }
if (!require('splines'))    { install.packages('splines');    library('splines') }

source("../utils.R")
source("../lab5/lab5_utils.R")

reads_file  <- "../lab2/data/TCGA-13-0723-10B_lib2_all_chr1.forward"
chr1_file   <- "../lab1/data/chr1_str.rda"

BIN_SIZE    <- 2500
beg_region  <- 50000000 
end_region  <- 100000000 

if (!file.exists(reads_file)) stop("File not found")
chr1_reads <- fread(reads_file, col.names = c("Chr", "Loc", "Length"))
load(chr1_file)

df_main <- bin_data(chr1_reads, chr1, beg_region, end_region, BIN_SIZE)

cat(sprintf("Loaded %d bins | GC range: %.3f-%.3f\n", 
            nrow(df_main), min(df_main$gc), max(df_main$gc)))
```

## Introduction

Lab 6 moves away from exploring the GC content \~ Reads phenomena, and utilizes the regression function we created in previous labs to investigate the goal we set out in the beginning of the semester - estimating the genomic copy number.
As we've learned in class, the genomic copy number is itself an indicator of possible mutations that has biological value, but is effected by additional technical effects.
Now that we've modeled the GC effect, we can try to estimate the distribution using our assumed model:

$$Y_i \sim \text{Pois}\left(a_i \cdot f(gc_i) + \epsilon_i\right)$$

For this lab, we use a healthy tissue sample as our baseline.
In healthy cells, we expect the copy number to be stable and close to 1, allowing us to establish a normal reference before eventually analyzing cancerous samples.

### Question 1

```{r}
estimate_copy_number <- function(reads_vec, gc_vec, fit_model) {
  # Predict expected coverage based on the GC spline model from lab4
  f_hat <- predict(fit_model, newdata = data.frame(gc = gc_vec))
  
  # Raw estimate: the ratio between observed and expected
  a_tilde <- reads_vec / f_hat
  
  # Normalize so the median equals 1 (the idenitifiablity assumption we had in class)
  m_hat <- 1 / median(a_tilde, na.rm = TRUE)
  return(a_tilde * m_hat)
}
```

To extract the copy number, we use the plug-in estimator presented in class:

$$\tilde{a}_k = \frac{y_k}{\hat{f}(gc_k)} \cdot \hat{M}$$

We also apply a normalization constant to ensure the median of our estimates is exactly 1 in order to align the data with the normalized baseline:

$$\hat{M} = \frac{1}{\text{med}_k\{\tilde{a}_k\}}$$

### Question 2

```{r, fig.height=9}
# fit GC-coverage spline (lab 4 model) with BIN_SIZE=2500
mod_spline <- fit_gc_spline(df_main)

# apply estimator to 1000 consecutive bins from the gene-rich region (61.25–63.75 Mb) we saw in lab5
df_sample <- df_main[4501:5500, ]
df_sample$a_hat <- estimate_copy_number(df_sample$reads, df_sample$gc, mod_spline)

# measure MAE from the healthy baseline of 1
mae_cn <- mean(abs(df_sample$a_hat - 1))
cat(sprintf("MAE from baseline: %.4f\n", mae_cn))

# visualize: marginal distribution and estimators along region
par(mfrow = c(2, 1))
hist(df_sample$a_hat, breaks = 30, col = "gray90", border = "white",
     main = "Marginal Distribution", xlab = "Estimated Copy number")
abline(v = 1, col = "red", lwd = 2, lty = 2)

plot(df_sample$a_hat, type = "l", col = "steelblue",
     main = "Copy number across Position", xlab = "Bin Index", ylab = "a_hat")
abline(h = 1, col = "red", lty = 2)
```

2a.
Visualizing the Estimates, The plots show that we successfully removed the GC bias.
The histogram is centered around 1, which is exactly what we expect for a healthy sample.
Looking at the spatial plot, the estimates are stable without any major jumps, giving us a more-or-less uniform copy number in the area we looked at.

\
2b: Measuring the MAE, we got \~0.15 (0.1443).
This means that each bin deviates from the baseline by 15% on average.
This deviation represents the residual noise left in our model: a mix of random sampling variation and other technical effects we mentioned in class as `seq`.

### Question 3

```{r, cache=TRUE}
# bin the full chromosome and filter out:
# - telomeres (<5 Mb, >242 Mb)
# - centromere (120–130 Mb)
# - bins with many N values, 0 reads
chr1_len   <- nchar(chr1)
df_full    <- bin_data(chr1_reads, chr1, 1, chr1_len, BIN_SIZE)
bin_pos_mb <- (seq_len(nrow(df_full)) * BIN_SIZE) / 1e6

keep   <- bin_pos_mb > 5 & bin_pos_mb < 242 &
          !(bin_pos_mb >= 120 & bin_pos_mb <= 130) &
          df_full$reads > 0
df_chr <- df_full[keep, ]

f_hat_chr <- suppressWarnings(predict(mod_spline, newdata = data.frame(gc = df_chr$gc)))

# Poisson part: Var[Y_k] = lambda_k  = a_k*f(gc_k) = 1*f(gc_k)
# So since we saw that Var_Y(a_k) = lambda_k / f^2(gc_k)
# => Var_pois = f(gc_k)/ f^2(gc_k)
var_pois <- f_hat_chr / f_hat_chr^2

# ε part: sigma^2 estimated empirically from spline residuals (no normality assumed)
sigma2  <- var(residuals(mod_spline))
var_eps <- sigma2 / f_hat_chr^2

mean_var_pois <- mean(var_pois)
mean_var_eps  <- mean(var_eps)
total_var     <- mean_var_pois + mean_var_eps

cat(sprintf("sigma^2:           %.5f\n", sigma2))
cat(sprintf("Poisson variance:  %.5f  (%.1f%%)\n", mean_var_pois, 100 * mean_var_pois / total_var))
cat(sprintf("epsilon variance:  %.5f  (%.1f%%)\n", mean_var_eps,  100 * mean_var_eps  / total_var))
cat(sprintf("Total variance:    %.5f\n", total_var))
```

The total estimated variance of $\hat{a}_k$ is `r round(total_var, 5)`, and comes from two sources - the inherent variance of the phenomena, modeled as the Poisson distribution, and the additional sequencing noise and technical effects.
The Poisson component contributes `r round(100 * mean_var_pois / total_var, 1)`% of it, and the $\epsilon$ component the other `r round(100 * mean_var_eps / total_var, 1)`%, giving us $\hat{\sigma}^2$ = `r round(sigma2, 4)`.

This means that most of the uncertainty in the model comes from the unexplained part - the additional sequencing-related technical effects.
So unless we try to model additional parts of the technical effects in this setup, the majority of the variance we see will remain.

### Question 4

In our original model, the noise is additive.
this suggests that we have small deviations in measurements that are both independent of the $f_{gc}$ and $a_k$ factors, and that the magnitude of the errors is independent of the magnitude of these factors.
So we should get some sort of random offset in our measurements.

The multiplicative model we see in this questions multiplies the factors $f_{gc}$ and $a_k$ by the error, so it effectively scales the magnitude of the error proportionally to these factors.
So here we get offset proportionate to the actual signal.

From the biological perspective, this means that the errors are generated by the sequencing process in a way that scales with amount of DNA we measure.

The new model does seem more fitting to our case, since the first model just assumes there's some external noise added to our final calculations.
We can view the second model as incorporating the noise proportionally to the NGS process output, and think of it as multiplying the noise in each sequencing stage.

## Short summary

This lab takes the new Poisson GLM model we built in class, and incorporates the GC estimator we built in previous labs, in order to first estimate the copy number.
We also estimate the variance and its components, and discuss the alternative multiplicative-error model.

## External sources statement

We used AI (Claude) to:

1.  rewrite the values in our explanation parts as inline r calculations
2.  rewrite the equations in the explanations as TEX
3.  add `fit_gc_spline()` to `lab5_utils.R` so the model specification is shared across labs
4.  fix the reads file path and add the missing `source("../lab5/lab5_utils.R")` call in the paths chunk
